### 2015-04-09 books/Logic_of_Political_Survival_Mesquita/code/ch41.tex
-

\documentclass[12pt,fleqn]{article}\usepackage{common}

\begin{document}

\begin{minted}[fontsize=\footnotesize]{python}
import pandas as pd
df = pd.read_csv('bdm2s2_nation_year_data_may2002.csv')
df['Klepto'] = (df['TAXGDP']-df['Expenditure']).abs()
df = df[['regyr','ccode','aid_gdp','lrgdpc','pop','year','W','S','Klepto','laglrgdpc']]
\end{minted}

\begin{minted}[fontsize=\footnotesize]{python}
import statsmodels.formula.api as smf
results = smf.ols('Klepto ~ W + S + aid_gdp + np.log(pop)', data=df).fit()
print results.summary()
\end{minted}

\begin{verbatim}
                            OLS Regression Results                            
==============================================================================
Dep. Variable:                 Klepto   R-squared:                       0.205
Model:                            OLS   Adj. R-squared:                  0.202
Method:                 Least Squares   F-statistic:                     65.82
Date:                Mon, 06 Apr 2015   Prob (F-statistic):           1.43e-49
Time:                        20:29:15   Log-Likelihood:                -3297.9
No. Observations:                1027   AIC:                             6606.
Df Residuals:                    1022   BIC:                             6630.
Df Model:                           4                                         
Covariance Type:            nonrobust                                         
===============================================================================
                  coef    std err          t      P>|t|      [95.0% Conf. Int.]
-------------------------------------------------------------------------------
Intercept       4.8089      1.223      3.933      0.000         2.410     7.208
W              -3.9128      0.879     -4.453      0.000        -5.637    -2.189
S               3.9783      0.675      5.892      0.000         2.653     5.303
aid_gdp        39.3581      2.858     13.770      0.000        33.750    44.967
np.log(pop)    -0.1965      0.115     -1.702      0.089        -0.423     0.030
==============================================================================
Omnibus:                      510.462   Durbin-Watson:                   0.657
Prob(Omnibus):                  0.000   Jarque-Bera (JB):             3672.715
Skew:                           2.179   Prob(JB):                         0.00
Kurtosis:                      11.176   Cond. No.                         141.
==============================================================================

\end{verbatim}

\begin{minted}[fontsize=\footnotesize]{python}
import statsmodels.formula.api as smf
results = smf.ols('lrgdpc ~ W + S + aid_gdp + np.log(pop) + laglrgdpc', data=df).fit()
print results.summary()
\end{minted}

\begin{verbatim}
                            OLS Regression Results                            
==============================================================================
Dep. Variable:                 lrgdpc   R-squared:                       0.993
Model:                            OLS   Adj. R-squared:                  0.993
Method:                 Least Squares   F-statistic:                 7.123e+04
Date:                Mon, 06 Apr 2015   Prob (F-statistic):               0.00
Time:                        22:08:39   Log-Likelihood:                 3221.6
No. Observations:                2541   AIC:                            -6431.
Df Residuals:                    2535   BIC:                            -6396.
Df Model:                           5                                         
Covariance Type:            nonrobust                                         
===============================================================================
                  coef    std err          t      P>|t|      [95.0% Conf. Int.]
-------------------------------------------------------------------------------
Intercept       0.0560      0.020      2.861      0.004         0.018     0.094
W               0.0262      0.007      3.992      0.000         0.013     0.039
S              -0.0086      0.004     -1.916      0.055        -0.017     0.000
aid_gdp        -0.0706      0.021     -3.339      0.001        -0.112    -0.029
np.log(pop)    -0.0012      0.001     -1.323      0.186        -0.003     0.001
laglrgdpc       0.9954      0.002    482.672      0.000         0.991     0.999
==============================================================================
Omnibus:                      476.268   Durbin-Watson:                   1.702
Prob(Omnibus):                  0.000   Jarque-Bera (JB):             4091.930
Skew:                          -0.635   Prob(JB):                         0.00
Kurtosis:                       9.086   Cond. No.                         221.
==============================================================================

\end{verbatim}

\begin{minted}[fontsize=\footnotesize]{python}
%load_ext rpy2.ipython
%R library(lme4)
\end{minted}

\begin{minted}[fontsize=\footnotesize]{python}
%R -i df
%R resp_lmer <- lmer(Klepto ~ W + S + log(pop) + aid_gdp + ( 1 | regyr), data = df)
%R -o res res = summary(resp_lmer)
print res
\end{minted}

\begin{verbatim}
Linear mixed model fit by REML ['lmerMod']
Formula: Klepto ~ W + S + log(pop) + aid_gdp + (1 | regyr)
   Data: df

REML criterion at convergence: 6569.8

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-1.9820 -0.6057 -0.1908  0.3338  7.5056

Random effects:
 Groups   Name        Variance Std.Dev.
 regyr    (Intercept)  5.245   2.290   
 Residual             32.296   5.683   
Number of obs: 1027, groups:  regyr, 122

Fixed effects:
            Estimate Std. Error t value
(Intercept)   6.1930     1.2722   4.868
W            -4.0960     0.8626  -4.748
S             3.5409     0.6541   5.414
log(pop)     -0.2682     0.1175  -2.284
aid_gdp      42.9981     2.9760  14.448

Correlation of Fixed Effects:
         (Intr) W      S      lg(pp)
W        -0.256                     
S        -0.171 -0.651              
log(pop) -0.903  0.195 -0.033       
aid_gdp  -0.417  0.071  0.043  0.329

\end{verbatim}

\begin{minted}[fontsize=\footnotesize]{python}
%R -i df
%R resp_lmer <- lmer(lrgdpc ~ W + S + log(pop) + aid_gdp + laglrgdpc + ( 1 | regyr), data = df)
%R -o res res = summary(resp_lmer)
print res
\end{minted}

\begin{verbatim}
Linear mixed model fit by REML ['lmerMod']
Formula: lrgdpc ~ W + S + log(pop) + aid_gdp + laglrgdpc + (1 | regyr)
   Data: df

REML criterion at convergence: -6453.3

Scaled residuals: 
    Min      1Q  Median      3Q     Max 
-9.6023 -0.4385  0.0367  0.5086  4.5498

Random effects:
 Groups   Name        Variance  Std.Dev.
 regyr    (Intercept) 0.0003723 0.01930 
 Residual             0.0042945 0.06553 
Number of obs: 2541, groups:  regyr, 217

Fixed effects:
              Estimate Std. Error t value
(Intercept)  0.0593032  0.0229343     2.6
W            0.0184700  0.0065761     2.8
S           -0.0042964  0.0043957    -1.0
log(pop)    -0.0017439  0.0009528    -1.8
aid_gdp     -0.0347754  0.0226003    -1.5
laglrgdpc    0.9957743  0.0023887   416.9

Correlation of Fixed Effects:
          (Intr) W      S      lg(pp) ad_gdp
W          0.152                            
S         -0.138 -0.644                     
log(pop)  -0.678  0.086 -0.044              
aid_gdp   -0.580 -0.058  0.044  0.453       
laglrgdpc -0.918 -0.296  0.122  0.370  0.472

\end{verbatim}

\end{document}

### 2015-07-23 dsync.sh
-

@@ -1,5 +1,5 @@
-disk="burak/New Volume"
-#disk="burak/17EA-3758"
+#disk="burak/New Volume"
+disk="burak/17EA-3758"
 echo "Copying files to /media/$disk/archive"
 rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/kitaplar "/media/$disk/archive"
 rsync  -W --size-only  --delete-after -P -avzh /home/burak/kod "/media/$disk/archive" 

### 2015-07-26 ms-thesis/ver1/biblio/bib.tex
-

\begin{thebibliography}{99}

\bibitem{gurgen} Gürgen, S. F., S. Yümlü., N. Okay,
  {\em A comparison of global, recurrent and smoothed-piecewise neural models
    for Istanbul stock exchange (ISE) prediction}, 
  Pattern Recognition Letters, Volume 26, Issue 13, Pages 2093-2103, 
  1 October 2005.

\bibitem{rabiner} Rabiner L. R. , 
  {\em A Tutorial on Hidden Markov Models and Selected Applications in Speech Recognition}, 
  Proceedings of IEEE vol. 77, no. 2, pp. 257-286, 
  1989.

\bibitem{ghahramani} Roweis S. and Z. Ghahramani,
  {\em  A Unifying Review of Linear Gaussian Models },
   Neural Computation 11(2):305--345, 
   1999.

\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
  {\em Parameter Estimation for Linear Dynamical Systems },
   Technical Report CRG-TR-96-2  
   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
   Department of Computer Science, University of Toronto, 
   1996.

\bibitem{alpaydin} Alpaydın E., 
  {\em Introduction to Machine Learning},
  MIT Press, 
  2004.

\bibitem{ghahramani3} Ghahramani Z., H. E. Hinton,
  {\em Switching State Space Models},
  Technical Report CRG-TR-96-3, Dept. Comp. Sci., Univ. Toronto, 
  1996.

\bibitem{shumway} Shumway R., H. S. Stoffer
  {\em Time series analysis and its applications 2nd Edition},
   New York, Springer, (Springer texts in statistics),
   2000.

\bibitem{jordan} Jordan M. I. , C. Bishop
  {\em An Introduction to Graphical Models},
   Not yet published, 
   2000.

\bibitem{kalman} Kalman R. E.,
  {\em A New Approach to Linear Filtering and Prediction Problems},
  Transactions of the ASME-Journal of Basic Engineering, 82 (Series D): 35-45, 
  1960.

\bibitem{kalman2} Kalman, R.E. and R.S. Bucy, 
  {\em New results in filtering and prediction theory},
  Trans. ASME J. Basic Eng., 83, 95-108,
  1961.

\bibitem{welling} Welling, M.,
  {\em The Kalman Filter - Lecture Tutorial},
   California Institute of Technology,
   2008.

\bibitem{lall} Lall, S.,
  {\em Modern Control 2 Lecture Notes},
   Stanford University,
   2006.

\bibitem{fama} Fama, E. F.,
  {\em The behavior of stock-market prices},
  The Journal of Business of the University of Chicago, 38, no. 1.
  1965.

\bibitem{famar} Fama, E. F.,
  {\em Random walks in stock market prices,},
  Financial Analysts Journal, 21 (5): 55-59.,
  1965.

\bibitem{tsay} Tsay, R. S.,
  {\em Time Series and Forecasting: Brief History and Future Research,},
  Journal of the American Statistical Association, Vol. 95, No. 450., pp. 638-643.,
  Jun 2000.

\bibitem{bachelier} Bachelier L.,
  {\em Thorie de la speculation. Annales Scientiques de l'Ecole Normale Superieure, III-17:21-86}
  Thesis for the Doctorate in Mathematical Science., 1900
  Reprinted by Editions Jacques Gabay, Paris 1995. English
  translation in The Random Character of Stock Market Prices, Ed. P. Cootner, pp
  1778, Cambridge, MIT Press, 1964.

\bibitem{shi} Shi S. and A. S. Weigend., {\em Taking Time Seriously: Hidden
  Markov Experts Applied to Financial Engineering}. In Proceedings of the
  IEEE/IAFE 1997 Conference on Computational Intelligence for Financial
  Engineering , pages 244-252. IEEE, 
  1997.

\bibitem{hassan} Hassan Md. R., Baikunth Nath, 
  {\em Stock Market Forecasting Using Hidden Markov Model: A New Approach},
  192-196, 
  ISDA 2005.

\bibitem{nabney} Azzouzi M.  and Ian T. Nabney,
  {\em Modelling Financial Time Series with Switching State Space Models},
  \PVerb!citeseer.ist.psu.edu/267552.html!,
  1999.

\bibitem{engle} Engle, Robert F. 
  {\em GARCH 101: The Use of ARCH/GARCH Models in Applied Econometrics},
  Journal of Economic Perspectives-Volume 15, Number 4-Pages 157-168, Fall
  2001.

\bibitem{engle2} Engle, R. F. , Victor K. Ng,  
  {\em Measuring and Testing the Impact of News on Volatility},
  The Journal of Finance, Volume XLVIII, No. 5, 
  December 1993.

\bibitem{engle3} Engle, Robert F., {\em Autoregressive Conditional
  Heteroskedasticity with Estimates of the Variance of United Kingdom
  Inflation}, Econometrica. 50:4, pp. 987-1007. University of California, San
  Diego. Revision,
  2002.

\bibitem{bollershev} Bollershev, T.,
  {\em Generalized autoregressive conditional heteroskedasdicity}, Journal of
  Econometrics 31, 307-327,
  1986.

\bibitem{nelson} Nelson D., 
  {\em Conditional heteroskedasticity in asset returns: A new approach},
  Econometrica, 59, 347-370, 
  1990.

\bibitem{mandelbrot} Mandelbrot B. B., 
  {\em The variation of certain speculative prices, Journal of Business},
  XXXVI (1963), pp. 392-417, 
  1963.

\end{thebibliography}

### 2015-07-26 ms-thesis/ver1/cover/ms-thesis-cover.tex
-

\title{FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN MARKOV MODELS}
\turkcebaslik{HMM VE KALMAN FİLTRELERİ KULLANARAK FİNANSAL ZAMAN SERİSİ TAHMİNİ}

\degree{B.E., Computer Engineering, Stevens Institute of Technology, 1996}

\author{Burak Bayramlı}

\program{Systems and Control Engineering}

\supervisor{Prof. Fikret Gürgen}

\examineri{Prof. Levent Akın}

\examinerii{Prof. Ethem Alpaydın}

\dateofapproval{..............................}

\begin{document}

\pagenumbering{roman}
\makemstitle        % For M.S. theses

\makeapprovalpage

\begin{acknowledgements}
I would like to thank my thesis advisor Dr. Gürgen for his help, direction and
patience during my work on this thesis.

Thanks to Dr. Ethem Alpaydın for first suggesting Systems and Control
Engineering department as a good master's goal, and second for all the knowledge
I gained in his {\em Pattern Recognition} class. Once in Systems and Control
Engineering department, I was able to take immensely interesting classes such as
{\em Nonlinear Dynamics} and {\em Simulation} both of which helped greatly in
researching this thesis. I am also grateful to Dr. Goldsman for the all
knowledge he was willing to part with in his {\em Simulation} class - I think
his lectures on pseudorandom generation will stay with me forever. It was also
in his class I first saw Brownian Motion which is very important in financial
time series analysis.

I extend warmest thanks to Dr. Levent Akın for his excellent {\em Autonomous
Robotics} class and his suggestion that I look into {\em Probabilistic
Robotics}, a book and term popularized by Sebastian Thrun - it was while
flipping through the pages of this book that I got the idea of using Kalman
Filters for stock price prediction.

I would also like to thank my parents Ercan and Füsun Bayramlı for their support
without whom none of this would have been possible.
\end{acknowledgements}

\begin{abstract}
Financial markets are challenging targets for analysis and prediction. The
existence of many factors that contribute to a final price of security at time
$t$, makes the task of predicting a future price a very hard task indeed. In the
past, many statistical and non-statistical models have been utilized that
attempted to perform this daunting task - the aim of this thesis is going to be
trying to demonstrate Hidden Markov Models and Kalman Filter methods can be used
for predicting a future price. We also devised a mixture predictor using HMM and
KF which we called KMM.

For this purpose, we hypothized HMM, KF and KMM based models that are trained on
historical data can generate future data, thereby predicting this securities'
price in the future. For data generation, we used Monte Carlo simulation to
smooth over the irregular patterns. ``Carrying the model forward in time'' is
achieved by a combination of Viterbi algorithm and rolling the dice on hidden
state transitions, in KF case, we follow the time transition equation.

Another goal of this thesis was to determine how HMM, KF and KMM methods stood
in comparison to other conventional methods. We picked polynomial regression,
Neural Networks (ANN) and plain ``random guess'' as our comparison
criteria. Random guess was expected to be the lowest performer in our tests,
every method mentioned should have surpassed random guess results by wide
margin. We were glad to see that this was indeed the case.
\end{abstract}

\begin{ozet}
Finans piyasaları, analiz ve fiyat tahmini yapmak açısından zor bir
alandır. Tahmin yapılmasını zorlaştıran faktör, herhangi bir gündeki fiyatın
ortaya çıkmasında çok fazla etkinin olmasıdır. Tahmin ve analiz için şimdiye
kadar pek çok istatistiki ve diğer yöntemler bu amaç için seferber edilmiştir -
biz de bu yolda bir adım atmaya uğraştık.

Bu bağlamda Gizli Markov Modelleri ve Kalman Filtreleri adlı iki kuvvetli
metotun, geçmiş senet verisini modelleyebilerek geleceğe dönük tahmin
yapabileceğini varsay\-dık, ve iki metodu birleştirerek KMM adlı üçüncü bir
metodun performansına baktık. Fiyat tahmin sayılarıni üretirken Monte Carlo
metotu kullanarak uç noktalarda olan tahminleri ortalama içinde düzeltmeye ve
böylece daha iyi bir tahmine ulaşmaya çabaladık. Model eğitildikten sonra onu
ileriye dönük tahminlerde kullanmak için ise, GMM durumunda Viterbi algoritması
ile geçiş olasılıklarını rasgele sayı üreterek takip eden bir algoritma, Kalman
Filtreleri durumunda ise en son gelinen zaman noktasından bir sonrakine geçiş
için bizim değiştirdiğimiz zaman güncelleme denklemleri, ve yine rasgele sayı
üreten bir algoritma kullandık.

Bu tezin bir diğer amacı tarif ettiğimiz metotların diğer klasik metotlara
kıyasla nasıl başarı göstereceğini ortaya koymaktı. Bunun için Yapay Sinir
Ağları, polinom regresyon ve en son pür rasgele sayı üreterek tahmin yapmaya
uğraşan üç diğer metotu test senaryolarımıza dahil ettik. Bu metotların arasında
rasgele üretim en alt seviyede performans göstermesini beklenilen metot idi,
diğer tüm metotlar farklı derecelerde bu metotu geçmeliydi, ki deneyler
sonucunda durumun böyle olduğunu gözlemledik.
\end{ozet}

\tableofcontents
\listoffigures
\listoftables

\begin{symabbreviations}
% The title will be typeset as "LIST OF SYMBOLS/ABBREVIATIONS".
% If you prefer it as "LIST OF SYMBOLS" or "LIST OF ABBREVIATIONS"
% use the environments "symbols" or "abbreviations", respectively.
%
% Use a separate \sym command for each symbols definition.
% First Latin symbols in alphabetical order

\sym{$a_{ij}$}{HMM transition probability $p(q_t^j=1 | q_t^i=1)$ of hidden states}
\sym{$A$}{HMM hidden transitions matrix. $a_{ij}$ accesses elements of $A$}
\sym{$A$}{Linear multiplier for KF time transition from time $t$ to $t+1$}
\sym{$C$}{A vector holding all means of emission Gaussians for HMM}
\sym{$C$}{Linear multiplier for KF that translates $x_t$  to $y_t$}
\sym{$h_t$}{Sum of errors going back $p$ days for ARCH / GARCH / EGARCH modeling.}
\sym{$K_t$}{Kalman gain matrix at time $t$} 
\sym{$P_t$}{Price of a security at time $t$}
\sym{$R_t$}{Rate of return for financial time series data}
\sym{$R$}{Variance of the emission Gaussian for HMM}
\sym{$P_{t}^{t}$}{Conditional variance of $x$  at time $t$ given output
emissions $y_0,...,y_t$}
\sym{$P_{t,t-1}$}{Joint covariance necessary for KF lag-one covariance smoothing
calculations} 
\sym{$q_t$}{HMM's hidden state symbol at time $t$}
\sym{$T$}{Number of training data points} 
\sym{$v_t$}{White noise - Gaussian distribution with zero mean and covariance $Q$}
\sym{$\hat{x}_{t}^t$}{Conditional expectation of $x$ at time $t$ given output
emissions $y_0,...,y_t$} 
\sym{$x_t$}{Random variable representing hidden state at time $t$ for KF
(Gaussian distribution).} 
\sym{$y$}{All of the measured/observed data for HMM, KF and KMM}
\sym{$w_t$}{White noise - Gaussian distribution with zero mean and covariance $R$}

\\\\

\sym{$\alpha(q_t)$}{Probability of emitting partial sequence of outputs
$y_0,...,y_t$ and ending up in state $q_t$ \cite{jordan} for HMM}
\sym{$\beta(q_t)$}{Probability of emitting partial sequence of outputs
$y_{t+1},..,y_T$ given that system starts in state $q_t$ for HMM}
\sym{$\eta$}{Distribution representing output emissions for HMM}
\sym{$\epsilon$}{Error}
\sym{$\gamma_{t}$}{HMM posterior probability $p(q_t|y)$ at time $t$}
\sym{$\pi$}{HMM initial probability distribution}
\sym{$\xi_t$}{HMM cooccurence probabilities of $q_t$ and $q_{t+1}$}

\\\\

\sym{ANN}{Artificial Neural Network}
\sym{ARCH}{Autoregressive conditional heteroskedasticity}
\sym{DJI}{Dow Jones Industrial}
\sym{EM}{Expectation Maximization}
\sym{GARCH}{General ARCH}
\sym{HMM}{Hidden Markov Models}
\sym{HME}{Hidden Markov Experts}
\sym{iid}{Independent Identically Distributed}
\sym{ISE}{Istanbul Stock Exchange}
\sym{KF}{Kalman Filters} 
\sym{KMM}{A hybrid method combining KF and HMM} 
\sym{MC}{Markov Chains} 
\sym{MLP}{Multilayer perceptron} 
\sym{RTS}{Rauch-Tung-Striebel} 
\sym{SSM}{State space model}

\end{symabbreviations}

### 2015-07-26 ms-thesis/ver1/hmm/ms-thesis-hmm.tex
-

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{HIDDEN MARKOV MODELS}

Hidden Markov Models are based on over hundred years old mathematical construct
called Markov Chains (MC). An MC can be in one of distinct $N$ states and the
transition between those states occur according to predefined transition
probabilities at discrete intervals. For the simplest, first-order Markov Chain,
these probabilities can be captured in a \PVerb!NxN!  matrix. The fact that
present state is enough to represent a system and is the only factor that
determines the future (through the transition table) is also called as the
system having the ``Markov property''. This also means that given present state,
the past and future are independent of eachother. Mathematically, a Markov Chain
is a sequence of random variables $X_1,X_2,X_3,...$, the state of the system at
time $t$ being $x_t$, then
\begin{eqnarray*}
P(X_{t+1} = x | X_t = x_t,...,X_1=x_1) = P(X_{t+1} = x | X_t = x_t)
\end{eqnarray*}
A Markov Chain could also be second order or third order and so on - for example
for second order Markov Chains, the transition table is \PVerb!2N x N! which
necessitates both present and {\em previous} step to obtain a transition probability
distribution for the next state.

Markov Chains have no hidden states. The emissions are the states themselves,
the transition probability matrix captures all the dynamics of the system. HMMs
improve on this model by adding a second layer of probability distribution that
represents observation values.

We assume a two level structure, with one level for hidden states, another for
observed states. Observed states model the ``emissions'', the output of the
model. In a typical setting where HMMs are used for prediction, modeler needs to
supply the number hidden states - an optimal value for this is usually
determined by using seperate training sets.

Hidden and observed states have different distributions.

\begin{figure}[!hbp]
\center{
  \scalebox{0.8}{
  \includegraphics{./hmm/hmm-intro.jpg}
  }
}
\caption{Hidden and observed states of an HMM}
\vspace{0.6cm}
\end{figure}

Following the style and derivations by Bishop and Jordan \cite{jordan}, hidden
states sequence which correspond to an observation sequence are represented as
$q_1,...,q_T$. The output sequence is represented by random variables
$y_1,...,y_T$ where $T$ represents the number of available data points. It will
be assumed that $q_t$ and $y_t$ are i.i.d $\forall t$ , that is, as time
progresses, the underlying distribution will not change for $q_t$ and
$y_t$. There is a conditional dependence between $y_t$ and $q_t$, $q_t$ and
$q_{t-1}$.

\begin{figure}[!hbp]
\center{
  \scalebox{0.8}{
  \includegraphics{./hmm/hmm-dep.jpg}
  }
}
\caption{Dependency between nodes of an HMM}
\vspace{0.6cm}
\end{figure}

We let random variable $q_t$ to have a multinomial distribution. There are $M$
finite states and for $q_t$ at state $i$, the``output'' of $q_t$ is a vector of
size M carrying a ``1'' at index i, ``0'' otherwise. Another representation that
can be used is $q_t^i = 1$ where $i \in \{1,..,M\}$ and $p(q_t)$ being a vector
of size $M$ where each cell carries the probability value for a possible value
of $q_t$.

There is conditional dependency between $y_t$ and $q_t$, and $q_t$ and
$q_{t-1}$. Since $q_t$ is multinomial, and there are $M$ states, joint
distribution between $q_t$ and $q_{t-1}$ should be represented by a table of
size $M$ x $M$. Let's call this $M$ x $M$ table the transition matrix
$A$. $a_{ij}$ gives transition probability $p(q_t^j=1 | q_t^i=1)$. Notice A is
independent of $t$.

Data output is captured in $y_t$. Output probability is $p(y_t|q_t)$. We make no
assumptions on this distribution for now.

One of our goals with HMM will be training it with ``given'' emissions (output)
$y_1,..,y_T$ to come up with matrix $A$ which can also be represented as
$p(q_t^j=1 | q_t^i=1)$. Another goal of HMM use is being able to compute a
likelihood - given a matrix $A$ and observation sequence $y$ we should be able
to compute the likelihood of a ``hidden'' state sequence $q$, more succintly
$p(q|y)$.

It's easier to compute $p(q_t|y)$ instead of entire $q$. We can use $p(y|q_t)$
to get to $p(q_t|y)$.
\begin{eqnarray*}
p(q_t|y) &=& \frac{p(y|q_t) p(q_t)}{p(y)}
\end{eqnarray*}
Then, we need to calculate $p(y|q_t)$ and $p(y)$. Trick: Conditioning rule
states given $q_t$,
\begin{itemize}
   \item $q_{t-1}$ becomes independent of $q_{t+1}$
   \item $y_{t-1}$ becomes independent of $y_{t+1}$.
\end{itemize}

% proofs were conditional indep were here see proofs.tex

\begin{eqnarray*}
p(q_t|y) &=& \frac{p(y|q_t) p(q_t)}{p(y)}\\
&=& \frac{p(y_0...y_t|q_t)p(y_{t+1}...y_T|q_t) p(q_t)}{p(y)}\\
&=& \frac{p(y_0...y_t,q_t)p(y_{t+1}...y_T|q_t)}{p(y)}\\
p(q_t|y) &=& \frac{\alpha(q_t)\beta(q_t)}{p(y)}
\end{eqnarray*}
where
\begin{eqnarray*}
\alpha(q_t) & \triangleq & p(y_0...y_t,q_t) \\
\beta(q_t) & \triangleq & p(y_{t+1}...y_T|q_t)
\end{eqnarray*}
Obtain $p(y)$ by applying $\sum_{q_t}$ on both sides
\begin{eqnarray}
\sum_{q_t}p(q_t|y)  &=& \sum_{q_t} \frac{\alpha(q_t)\beta(q_t)}{p(y)} \nonumber\\
1  &=& \sum_{q_t} \frac{\alpha(q_t)\beta(q_t)}{p(y)} \nonumber\\
p(y) &=& \sum_{q_t} \alpha(q_t)\beta(q_t) \label{py}
\end{eqnarray}

\section{Forward-Backward Recursion}

We are finished with $p(y)$. We can move on to $\alpha(q_t)$ and
$\beta(q_t)$. We will compute these recursively. That is, from $\alpha(q_t)$ we
will compute $\alpha(q_{t+1})$.

\begin{eqnarray*}
\alpha(q_{t+1}) &=& p(y_0,...,y_{t+1},q_{t+1}) \\
 &=& p(y_0,...,y_{t+1}|q_{t+1})p(q_{t+1})\\
 &=& p(y_0,...,y_t|q_{t+1})p(y_{t+1}|q_{t+1})p(q_{t+1})\\
 &=& p(y_0,...,y_t,q_{t+1})p(y_{t+1}|q_{t+1})\\
 &=& \sum_{q_t}p(y_0,...,y_t,q_t,q_{t+1})p(y_{t+1}|q_{t+1})\\
 &=& \sum_{q_t}p(y_0,...,y_t,q_{t+1}|q_t)p(q_t)p(y_{t+1}|q_{t+1})\\
 &=& \sum_{q_t}p(y_0,...,y_t|q_t)p(q_{t+1}|q_t)p(q_t)p(y_{t+1}|q_{t+1})\\
 &=& \sum_{q_t}p(y_0,...,y_t,q_t)p(q_{t+1}|q_t)p(y_{t+1}|q_{t+1})\\
 &=& \sum_{q_t}\alpha(q_t)a_{q_t,q_{t+1}}p(y_{t+1}|q_{t+1})
\end{eqnarray*}
Similarly we can derive the equation for $\beta(q_t)$.
\begin{eqnarray*}
\beta(q_t) &=& p(y_{t+1},...,y_T|q_t) \\
&=& \sum_{q_{t+1}} p(y_{t+1},...,y_T,q_{t+1}|q_t) \\
&=& \sum_{q_{t+1}} \frac{p(y_{t+1},...,y_T,q_{t+1},q_t)}{p(q_t)}\\
&=& \sum_{q_{t+1}} \frac{p(y_{t+1},...,y_T,q_t|q_{t+1})p(q_{t+1})}{p(q_t)}\\
&=& \sum_{q_{t+1}} p(y_{t+1},...,y_T,q_t|q_{t+1})p(q_{t+1}|q_t)\\
&=& \sum_{q_{t+1}} p(y_{t+2},...,y_T,q_t|q_{t+1})p(q_{t+1}|q_t)p(y_{t+1}|q_{t+1})\\
&=& \sum_{q_{t+1}} \beta(q_{t+1}) a_{q_t,q_{t+1}} p(y_{t+1}|q_{t+1})
\end{eqnarray*}

We see that
\begin{itemize}
   \item $\alpha$ recursion is a forward recursion (we plug in $q_t$ to get
     $q_{t+1}$)
   \item $\beta$ recursion is a backward recursion (we plug in $q_{t+1}$ to get
     $q_{t}$)
\end{itemize}
The start-up equation can be calculated as follows:
\begin{eqnarray*}
\alpha({q_0}) &=& p(y_0,q_0)\\
&=& p(y_0|q_0)p(q_0)\\
&=& p(y_0|q_0)\pi_0
\end{eqnarray*}
There is a little complication for the start-up formula for $\beta$
recursion. In this case it would not be possible to start with $\beta(q_{T+1})$
because that state does not exist in our system. In this case, we can simply
start with vector of ones. For the proof the reader can refer to \cite{jordan}.

% proof 2 buradaydi

% EM param estimation buradaydi
\section{Training}

We need to find optimal $\theta$ that maximizes $log (p(y,q|\theta))$ where
$\theta = (\pi,A,\eta)$. We know $\pi$ and $A$. $\eta$ represents parameters
necessary for $p(y_t|q_t)$, the output distribution.

Let's write the full $p(y,q)$ distribution;
\begin{eqnarray*}
p(q,y) &=& \pi_{q_0} \prod_{t=0}^{T-1} a_{q_t,q_{t+1}} \prod_{t=0}' p(y_t|q_t,\eta)
\end{eqnarray*}
We take log, use Lagrange multipliers, take derivative set to zero and solve:
{\small
\begin{eqnarray*}
\hat{a}_{ij}  &=& \frac{\sum_{t=0}^{T-1} q_t^i q_{t+1}^j}{\sum_{j=1}^M \sum_{t=0}^{T-1} q_t^i q_{t+1}^j} 
\end{eqnarray*}}
Similarly for $\eta_{ij}$ and $\pi_i$,
\begin{eqnarray*}
\hat{\eta}_{ij} &=& \frac{\sum_{t=0}^{T-1} q_t^i y_t^j}{\sum_{k=1}^N
\sum_{t=0}^{T-1} q_t^i y_t^k} \\
\hat{\pi}_i &=& q_0^i
\end{eqnarray*}

$\hat{\eta}_{ij}$ assumes a multinomial output distribution $y_t$. For stock
prediction purposes, we will be using a Gaussian distribution to represent
each stock price at time $t$. Therefore, we can simply use $y_t$ itself as the
output instead of the $y_t^j$, hence $\hat{\eta}$ changes to being the mean of
the estimated Gaussian.
\begin{eqnarray*}
\hat{C}_{i}^{(p+1)} &=& \frac{\sum_{t=0}^T\gamma_{t}^i y_t}{\sum_{t=0}^T\gamma_{t}^i}
\end{eqnarray*}
where $\gamma_{t}$ is the posterior probability $p(q_t|y)$ at time $t$.

In each iteration

\begin{itemize}
   \item We calculate $\hat{a}$, $\hat{C}$, $\hat{R}$, $\hat{\pi}$ using given
     $q$ and $y$.
   \item We calculate $p(q_t|y)$ for each $q_t$ using $A$, $\eta$ and $\pi$ using
   forward backward recursion.
\end{itemize}

\section{Algorithm}

The algorithm for HMM forward-backward inference and training through EM
(Baum-Welch) as presented by Roweis and Ghahramani \cite{ghahramani} can be
found below. The operator $\odot$ is defined to be element-wise product of two
equal size vectors or matrices. The algorithm assumes the emission distribution
is a multidimensional Gaussian.

\begin{figure}[!hbp]
\center{
\begin{tabbing}
 {\tt HMM}\={\tt Inference}\=(Y,A,$\pi$,C,R) \\
 {\tt for} t=1  to  T  \% forward pass \\
 \> {\tt for} i=1  to k \\
 \> \>$b_{t,i}=\texttt{exp}\{-\frac{1}{2}(y_t - C_i)'R^{-1}(y_t - C_i)\}|R|^{-1/2}(2\pi)^{-p/2}$\\
 \> {\tt end}\\
 \> if (t=1) $\alpha_t = \pi \odot b_1$ else $\alpha_t = [A'\alpha_{t-1}]\odot b_t$ end \\
 \> $\rho_t = \sum_i \alpha_{t,i}$\\
 \> $\alpha_t = \alpha_t / \rho_t$ \% normalization\\
 {\tt end}\\
 $\beta_T = 1/\rho_t$ \\
 {\tt for }t=T-1 to 1 \% backward pass \\
 \> $\beta_t = A[\beta_{t+1} \odot b_{t+1}]/ \rho_t$\\
 {\tt end}\\
 $\gamma_t = \alpha_t \odot \beta_t / (\alpha_t'\beta_t)$\\
 $\xi_t = \alpha_t[\beta_{t+1} \odot b_{t+1}]'$\\
 $\xi_t = \xi_t / (\sum_{ij}\xi_{ij})$\\
 {\tt return} $\gamma_t, \xi_t, \rho_t$ for all t
\end{tabbing}
}
\caption{HMM Inference Algorithm}
\vspace{0.6cm}
\end{figure}

\begin{figure}[!hbp]
\center{
\begin{tabbing}
{\tt HMM}\={\tt Lea}\={\tt rn}(Y,k,$\epsilon$) \\
\> {\tt initialize}  A, $\pi$, C, R \\
\> {\tt while change in log likelihood} $>$ $\epsilon$ \\
\> \> {\tt HMMInference}$(Y,A,\pi,C,R);$ \%E step\\
\> \> \% M Step\\
\> \> $\pi = \gamma_1$\\
\> \> $A = \sum_t \xi_t$\\
\> \> $A_{ij} = A_{ij} / \sum_l A_{il}$ {\tt for all }$ i,j$\\
\> \> $C_j = \sum_t \gamma_{t,j} y_t / \sum_t \gamma_{t,j}$ {\tt for all } $j$\\
\> \> $R = \sum_{t,j} \gamma_{t,j} (y_t - C_j)(y_t - C_j)' / T$\\
\> {\tt return } $A, \pi, C, R$
\end{tabbing}
}
\caption{HMM Learning Algorithm}
\vspace{0.6cm}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Prediction with HMM}

For HMMs, if we start generating using $\pi$, this would give us a replay of the
training data starting with day $1$, not future data starting at $T+1$. Then, we
should not start generating starting with $\pi$, instead, we need to determine
{\em the last hidden state of our HMM} on day $T$, and use transition table $A$
to get to hidden state at day $T+1$, {\em then} we can start generating predictions.

In order to find the hidden state at day $T$, we can simply run the Viterbi
algorithm that gives us the most likely hidden path for the given data and given
HMM. Then, we simply pick the last element from this list, roll the dice and
apply transition once, and this gives us our first prediction for day $T+1$. The
rest will be the same as generating data from a regular HMM - we keep following
the transition table. The pseucode for the function
\PVerb!observationGeneration!  looks as follows:

\begin{figure}[!hbp]
\center{
\begin{tabbing}
{\tt obser}\={\tt vation}\={\tt Genera}\={\tt tion}$(A,\pi,C,R,N)$\\
\> $states$ = {\tt viterbi}($A,\pi,C,R$);\\
\> $lastState = states[T]$;\\
\> {\tt } for $i = 1$ {\tt to } $N$\\
\> \> $currentState = {\tt nextState}(lastState,A,\pi,C,R);$\\
\> \> $mean = C[currentState]$;\\
\> \> $variance = R[currentState];$\\
\> \> $o[i] = $ {\tt gaussianRandomGeneration}$(mean,variance);$\\
\> \> $lastState = currentState;$\\
\> {\tt end}\\
\> {\tt return} $o$;
\end{tabbing}
}
\caption{Algorithm for Generating Observations}
\vspace{0.6cm}
\end{figure}

\newpage

\begin{figure}[ht]
\center{
\begin{tabbing}
{\tt next}\={\tt Sta}\={\tt te}$(stat$\=$e,A,\pi,C,R);$\\ 
\> $randVal = ${\tt generateUniform(0,1);}\\
\> $total = 0;$\\
\> $done = false;$\\
\> $total = A[currentState][0];$\\
\> {\tt for } $j = 1$ {\tt to } $numStates$\\
\> \> {\tt if} $done = true$ {\tt break};\\
\> \> {\tt if} $(randVal > total$ {\tt and } \\
\> \> \> $randVal \leqslant (total + A[currentState][j]))$\\
\> \> \> $currentState = j;$\\
\> \> \> $done = true;$\\
\> \> {\tt end}\\
\> \> $total = total+ A[currentState][j]$;\\
\> {\tt end}\\
\> {\tt return} $currentState$;
\end{tabbing}
}
\caption{State Transition Algorithm}
\vspace{0.6cm}
\end{figure}

### 2015-07-26 ms-thesis/ver1/intro/ms-thesis-intro.tex
-

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{INTRODUCTION}

The aim of this thesis is to explain and apply Hidden Markov Model (HMM) and
Kalman Filter (KF) methods in context of time series prediction
problem. Istanbul Stock Exchange (ISE) and Dow Jones Industrial (DJI) index
historical prices were used as a testbed for both models and algorithms. In
order to show the capability of HMM, KF and KMM methods, various other
non-statistical methods were also used on the same dataset among which are ANNs,
polynomial fitting, and finally a simple random number generator based
predictor.

We also combined HMM, KF and KMM methods in a hybrid mechanism called KMM. This
method tries to merge both HMM, KF and KMM results through predefined weights. It was
hoped that the total would be better than the sum of its parts, that is, HMM and
KF combined could yield better and different results than each method taken
seperately.

ISE and DJI data sets consisted of daily closing prices for over 21 years for
ISE and 30 years for DJI. DJI data was obtained from
\PVerb!http://finance.yahoo.com! site. ISE-100 historical data is available from
Turkish Central Bank site at \PVerb!http://evds.tcmb.gov.tr!. We attempted to
predict future price for 1, 5, and 10 days for each method mentioned above. All
methods were given a limited ``time window'' on which training took place, after
which, prediction for future prices could be performed. We kept the training
window the same size for all algorithms with the exception of ANN because too
much data was causing overfitting and effecting results negatively.

A natural benchmark for success of HMM, KF, KMM algorithms, or any predictive
algorithm, should be it's relative success against a random predictor. This idea
was inspired by Fama \cite{famar} where he states ``if [an] analyst can make
meaningful judgments concerning the purchase and sale of individual securities,
his choices should consistently outperform {\em randomly} selected securities of
the same general riskiness''. One can also say the worst-case performance, lower
bound of any financial predictive algorithm should be random prediction. Any
algorithm which can beat random prediction can claim that it is calculating
``some'' trend in the market that ``pure chance'' does not.

Before further delving into the details of each algorithm, we will briefly
introduce time series analysis in general, and financial time series analysis in
particular. We will provide a motivation for the methods we used, the details of
HMM, KF and KMM methods, the results of all algorithms mentioned and finally an
analysis of the outcome.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Time Series}

Time series data analysis is a natural by-product in many fields as diverse as
seismology, biology, physics, speech processing, medicine and more. Such fields
produce experimental data that needs to be analyzed using specific time series
methods and in certain cases, this analysis needs to be conducted through
probabilistic, statistical models.

Historically time series analysis was divided into two schools: {\em Frequency
  domain} and {\em time domain} approaches. Time domain approaches stated
current data was a function of earlier data and concocted an autocorrelation
function $\rho_l$ of the data, frequency approaches used spectral analysis where
Fourier transform of the autocorrelation function became important
\cite{tsay}. The sharp divisions that seemed to exist between the two approaches
seem to be settled, and at the same time the use of state space models and
Kalman filtering in 1980s caused an explosion of interest in related models for
time series analysis.

Statistical models try to explain sample data using a collection of random
variables that are in order, that fluctuate over time. For example say a time
series sample is represented by a collection of 10 random variables
$x_1,..,x_{10}$ where random variable $x_1$ represents the value of the series
at the first time point, $x_2$ at the second time point, so on. A collection of
random variables ${x_t}$ indexed by $t$ is also referred to as a stochastic
process \cite{shumway}. An excellent review of time series analysis methods can
be found in \cite{tsay}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Financial Time Series}

Financial time series data brings forth certain issues with it from a
statistical analysis standpoint, most important of which is the interdependency
of one data point with previous data points. As a result of this
interdependency, the modeler of the data is unable to assume each point to be
independent and identically distributed (i.i.d.) which complicates the modeling
process.

The collapse of Long-Term Capital Management in 1998 proved to modelers of
financial time series that one could not see fluctuations of stock prices as
though coming from i.i.d. normal density functions. The said company based its
investment on such an model, and when it collapsed the situation threatened to
open a trillion-dollar hole in the banking industry. Since Bachelier
\cite{bachelier}, constructing finance models on Brownian Motion had been
popular until scientists started to poke holes in the theory and real-life
catastrophes such as the one above proved otherwise. Brownian Motion based
modeling requires that at each time step the price difference was a result of an
indepedent normal distribution. Fama \cite{fama} offered a detailed explanation
of the problem with that approach:

\begin{quote}
``If the population of price changes is strictly normal, on the average for any
stock ... an observation more than five standard deviations from the mean
should be observed about once every 7,000 years. In fact such observations
seem to occur about once every three to four years.''
\end{quote}

In other words Fama's findings indicated that independence assumption was
incorrect. HMM and KF methods construct a Bayesian graphical model that tries to
address this problem.

\begin{figure}[ht]
\center{
  \scalebox{0.30}{
  \includegraphics{./intro/somevals.jpg}
  }
}
\caption{Some Values from DJI and ISE-100}
\vspace{0.6cm}
\end{figure}

\section{Return of Financial Time Series}

We wanted to limit variations in price changes to a respectable range of values,
therefore chose to utilize rate of return of ISE and DJI price changes, instead
of the absolute values themselves. For time $t$, the return $R_t$ can be
calculated as
\begin{eqnarray*}
R_t = \frac{P_t - P_{t-1}}{P_{t-1}}
\end{eqnarray*}
where $P_t$ is the price of the financial instrument at time $t$.

In order for HMM, KF and KMM models to learn from the data, we first pre-processed it
converting regular prices into rate of returns. This way after training and
prediction, generated values from both models would automatically be daily
returns themselves. However if an application required predictions in the form
of ``absolute stock values'', one needs go from rate of returns back to absolute
values, in this case one can simply take last absolute value for known price of
that security, and start multiplying with the predicted rate of returns
values. For example, if training was performed on $P_1,...,P_T$ where $T$ is the
amount of training points, then HMM or KF would predict returns for
$R_{T+1},...,R_{T+30}$, and using the last known price point $P_T$, we would
start multiplying our generated returns and obtain $P_{T+1},...,P_{T+30}$.

\begin{figure}[!hbp]
\center{
  \scalebox{0.30}{
  \includegraphics{./intro/volimkb.jpg}
  }
}
\caption{IMKB Volatility}
\vspace{0.6cm}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Istanbul Stock Exchange and Dow Jones Industrial Data}

Now let us look at the data more closely. ISE National-100 index and Dow Jones
Industrial index are not individual stock prices themselves, instead they are
{\em summaries} over certain stock prices. This summarization capability makes
them good indicators, as economists like to say, especially an index such as
ISE-100 can signal good or bad times in the nation's stock market because of the
many number of companies' stocks it choses to reflect.

There are many methods to calculate an index. Some methods calculate a simple
average, e.g. Dow Jones Industrial index is such an average. 30 industrial
companies with high enough valuation in New York Stock Exchange were picked to
form this index, and the stock price for all 30 companies were simply averaged
to obtain a final DJI value.

ISE National-100 index on the other hand, is an index that is a weighted average
- it takes more than one parameter to calculate a final result and it assigns
different weights to each financial instrument. The formula is
\begin{eqnarray*}
E_t = \frac{\sum_{i=1}^n F_{it} N_{it} H_{it}}{B_t}
\end{eqnarray*}
where\\
$E_t$: The value of index at time $t$\\
$n$: The number of companies that contribute to the index' calculation\\
$F_{it}$: The value of $i^{th}$ stock at time $t$\\
$N_{it}$: The total number of shares outstanding of the stock $i$ at time $t$. \\
$H_{it}$: The flotation weight (publicly-held portion, i.e. the ratio of stocks
kept in custody at Takasbank, except those kept in non-fungible accounts) of the
stock $i$ at period $t$. \\
$B_t$: The value of divisor at period $t$ (Adjusted base market value). \\

Here are some statistics on the DJI and ISE data:

\vspace{0.3cm}

\begin{table}[!h]
\caption{Istanbul Stock Exchange}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
Size & Mean & Var & Kurtosis & Skew & J.Bera Null & Min & Max \\
\hline
5045 & 0.00214 & 0.0008 &     6.9154 &     0.2745 &  1 & -0.1811 & 0.242 \\
\hline
\end{tabular}
\end{table}

\vspace{0.3cm}

\begin{table}[!h]
\caption{Dow Jones Industrial}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|l|}
\hline
Size & Mean & Var & Kurtosis & Skew & J.Bera Null & Min & Max \\
\hline
9341 & 0.00035 & 0.0001 &    34.40 &    -1.15 &     1 &    -0.226 & 0.101 \\
\hline
\end{tabular}
\end{table}

Notable statistics presented above are kurtosis, skew and Jarque-Bera
test. Skewness is a measure of asymmetry, visually a distribution with positive
skewness would have a so-called tail much longer on its right side than on its
left. The reverse statement is true for negative skewness - such distributions
would have a longer tail on the left side of its graphic. Longer tail on one
side means more data would come from that side of that distribution.

\begin{figure}[!h]
\center{
  \scalebox{0.55}{
  \includegraphics{./intro/skew.JPG}
  }
}
\vspace{0.6cm}
\end{figure}

Mathematically, skewness is the third standardized moment $\gamma_1$ calculated
as 
\begin{eqnarray*}
\gamma_1 = \frac{\mu_3}{\sigma^3}
\end{eqnarray*}
where $\mu_3$ is the third moment and $\sigma$ is the well-known standard
deviation. Skewness can signal us if one can assume a distribution is
Gaussian. Gaussian distribution has a skewness of zero, hence if sample skewness
of a distribution is non-zero, we can conclude that this distribution is not
Normal.

Kurtosis, another important metric, tells us something about the presence of
extremely high values in the data. Standard normal distribution has a kurtosis
of zero, higher kurtosis values signal the presence of sharper peak in the
distribution, lower kursosis distributions tend to have flatter peaks. In other
words, kurtosis metric can tell us if extreme values are found more commonly in
a distribution or not. The formula for kurtosis is 
\begin{eqnarray*}
\frac{\sum_{i=1}T(X_i - \bar{X})^4}{(T-1)\sigma^4}
\end{eqnarray*}
Taken together skewness and kurtosis metrics can indicate if a distribution is
normal. There is also a goodness-of-fit measure called Jarque-Bera test that can
indicate normalty of a distribution. This test combines skewness and kurtosis
measures to accept or reject the null hypothesis that the distribution in
question is normal.

Looking at skewness and kurtosis for ISE and DJI data sets, we can see kurtosis
value is very high for both of the sets. This means there are a lot of extreme
values present in the sample data. It can also be observed that ISE-100 is
skewed positively, though not by much, and DJI data is skewed negatively. It is
known in finance literature that stock markets are skewed negatively because
decreases are much sharper than increases. An hypothesis as to why claims that
the reason is related to basic human psychology: At time of crises, people panic
and group mentality tempts a big number of people to sell their stocks at the
same time, hence causing sharper decreases than increases in the stock market
data. The claim seems to be justified by the presence of certain ``brakes'' in
the market in almost all markets in the world with the help of which, if daily
market decrease passes a certain predefined threshold, all activity in the
market is stopped until the next day. The goal here is giving traders time to
think and calm down, trying to offset the selling histeria that might be caused
in a moment of confusion.

We mentioned almost all markets are skewed negatively, then we need to explain
why ISE-100 skewed positively. The reason here could be that ISE is a relatively
young market and the composition of ISE-100 index has changed over the years,
hence the sample skewness for 20 years worth of data points (as we have done)
might not offer the best results. We hypothized taking later years' data points
would offer negative skewness. This turned out to be correct, a skewness
calculation for around last 8 years did indeed offer a negative result.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Classic Methods: ARCH, GARCH, EGARCH models}

Least squares models have been used on financial time series data for many
years. One of the important assumptions that least squares based analysis makes
is the variance of error terms stays the same throughout the time series. This
assumption, which is called homoskedasticity, does not seem to hold for
financial time series data consistently. It has been shown that the reverse
condition is true, called heteroskedasticity, that posits the variance of the
error changes. ARCH/GARCH methods attempt to model this variance \cite{engle}.

ARCH stands for \textbf{a}uto\textbf{r}egressive \textbf{c}onditional
\textbf{h}eteroskedasticity, and GARCH is \textbf{g}eneral ARCH. First proposed
by Engle \cite{engle3} then Bollershev \cite{bollershev}, these models try to
put forth a correlation between variance of error at time $t$ and variance on
days preceding $t$. The reason for such relationship is the fact that volatility
seems to have ``regions'' as observed by Mandelbrot in \cite{mandelbrot} - he
stated that in rate of return data ``large changes tend to be followed by large
changes, of either sign, and small changes tend to be followed by small
changes''. This and similar observations led to an array of research on
volatility clustering using methods such as ANN and many other statistical
tools. (G)ARCH methods are one of these methods which try to forecast volatility
based on historical rate of return data. As of 1992, a survey showed over 200
papers had used (G)ARCH and related models for financial time series
\cite{engle2}.

As shown in \cite{engle2}, let $r_t$ be the rate of return of a stock from time
$t-1$ to $t$, and $F_{t-1}$ be ``information'' that is relevant to make a
decision about buying or selling a security at time $t-1$. Then the expected
rate of return at time $t$ can be shown as $E(r_t|F_{t-1})$ and the variance can
be shown as $Var(r_t|F_{t-1})$. Let us call the former $m_t$ and the latter
$h_t$. Engle \cite{engle3} suggests that $h_t$ can be modeled as a function of
$\epsilon_t$ which is the error calculated as $\epsilon_t = r_t - m_t$. In a
ARCH(p) model this can be shown as 
\begin{eqnarray*}
h_t = \omega + \sum_{i=1}^p \alpha_i\epsilon_{t-i}^2
\end{eqnarray*}

where $\omega$, $\alpha_1,..,\alpha_p$ and $p$ are all constants which (except
$p$) are to be estimated from the past historical data. According to this
ARCH(p) model, past errors (news) older than $p$ periods have no effect on the
current day.

GARCH(p,q) model invented by Bollershev \cite{bollershev} generalizes ARCH(p)
model as
\begin{eqnarray*}
h_t = \omega + \sum_{i=1}^p \alpha_i \omega_{t-1}^2 + \sum_{i=1}^q \beta_i h_{t-i}
\end{eqnarray*}
where $\alpha_1,..,\alpha_p,\beta_1,...,\beta_p$ and $\omega$ are constants.

EGARCH model proposed by Nelson \cite{nelson} attempts to correct problems in
(G)ARCH modeling that stem from the asymetric nature of the stock price data:
Statistical analysis of volatility showed that drops in markets are sharper and
deeper than increases as we have seen previously. In comparison good news in the
market is taken in much more rationally and causes more moderate increases than
decreases. (G)ARCH models do not take this asymetry into account which EGARCH
models attempt to correct. EGARCH is formulized as
\begin{eqnarray*}
log(h_t) = \omega + \beta \cdot log(h_{t-1}) + 
\gamma \frac{\epsilon_{t-1}}{\sqrt{h_{t-1}}} + 
\alpha \bigg[ \frac{|\epsilon_{t-1}|}{\sqrt{h_{t-1}}} - \sqrt{2 / \pi} \bigg]
\end{eqnarray*}
where $\omega,\beta,\gamma$ and $\alpha$ are constants. As a result of this
formulation positive $\epsilon$ values generate less volatility than negative
values - or good news have less effect than bad news. Mathematically, out of two
$\epsilon$ values that have the same absolute value but different signs, the
negative one would cause more increase. Notice how $\gamma$ in $\gamma \cdot
\epsilon_{t-1} / \sqrt{h_{t-1}}$ can now capture the sign of earlier $\epsilon$
values, whereas in ARCH and GARCH models this term was squared causing only the
amplitude of $\epsilon$ to contribute to the model, whether it was negative or
positive. Also, EGARCH model assures that ``bigger'' news have ``greater''
impact on volatility which GARCH model does not take into account \cite{engle2}.

All ARCH variants with preference being on GARCH and EGARCH have been found
quite successful in predicting market volatility. We did not use them in our
comparison of prediction methods.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Artificial Neural Networks}

ANNs are biologically inspired constructs which are used heavily in robotics,
time series analysis, signal filtration, regression and classification. ANNs are
built to mimic neurons in human brain which can receive signals through
dendrites, and emits them through axons to snyapses, then the signal is
transmitted to the next neuron in the chain.

\begin{figure}[h]
\center{
  \scalebox{0.8}{
  \includegraphics{./intro/neuron.JPG}
  }
}
\caption{A neuron}
\vspace{0.6cm}
\end{figure}

A network with one layer of neurons is called a perceptron. If there is more
than one layer available, this network is called a multilayer perceptron
(MLP). For the purposes of this thesis, we used an MLP for our time series
analysis tasks. It has been proven that an MLP with single hidden layer and
sufficient number of neurons can approximate any function. This is an
improvement over single layer regular perceptron which has problems with the
infamous XOR function.

ANNs are trained with using data sets whose contents include a set of input
points $x_1,..,x_N$ along with output data point(s) $y_1,..,y_K$. At each
training step, the current state (weights of the neurons) of ANN is used to
calculate an error between predicted and real output values, then, the weights
of the hidden neurons are modified in order to decrease the error value. This is
performed for all data points in the training data set. The error measure $J$ is
\begin{eqnarray*}
J = \frac{1}{2N}\sum_{i=1}^n (\hat{y}_i - y_i)^2
\end{eqnarray*}
Gradient descent with back propagation is a popular method to minimize the error
function at each iteration.

For stock price prediction purposes, we utilized an MLP network with 30 neurons
in the hidden layer, input size 300, one output and linear activation
function. We used a Matlab\textsuperscript{\textregistered} package called
Netlab as our ANN package written by Christopher Bishop and Ian Nabney.

Time series data had to be tailored so it could be fed into a neural network. In
applications such as face recognition, flattening out the input matrix into one
continous vector $\mathbf{x}$ and mapping this flattened structure to a single
classification code is quite natural. With time series, data the mapping of
input to outputs is somewhat less intuitive. The trick is using autoregression -
having the time series contain both the input and the output at the same time
where one data input set $x_1,...,x_N$ can act as input and $x_{N+1}$ can act as
output, next input would be $x_2,...,x_{N+1}$ as input and $x_{N+2}$ as output
and by shifting by one, so on. With this method, at each iteration we
essentially are creating a new dataset which contains a set of input/output
matchings.

To make predictions, we feed ``current data'' starting from today going back as
much as input allows, as a new input collection, and the single output value is
then the prediction for tomorrow's stock price. If we need to make prediction
for more than one day, we treat each newly generated predicted value as another
input, folding it into the same time series sequence as the rest of the input
data. Naturally this would cause predictions to worsen as part of the input data
starts consisting of not actual data but predicted data, but if the modeler
keeps the projected days to a minimum, the algorithm can produce acceptable
results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Polynomial Regression}

Another simple yet powerful method used for comparion was a method that falls
under linear regression umbrella, called polynomial regression. With this model,
given data for $y$ and $x$ pairs, the relation between them is assumed to be

\begin{eqnarray*}
y = \beta_0 + \beta_1x + ... + \beta_1x^k + \epsilon
\end{eqnarray*}

In order to fit the model to the data, the modeler first decides the order of
the polynomial equation (we used 10). Then, a least squares fit is performed
which aims to minimize the error $\epsilon$. At the end this process outputs all
$\beta$ coeffecients to define the equation for $y$. Matlab's \PVerb!polyfit!
accomplished this task.

During modeling process, we first tried ``time'' variable as $x$ and obtained
less than satisfactory results. Next we decided to divide up the historical
stock price data in two parts, and have $i^{th}$ element of first part to be
$x$, and $i^{th}$ element of second part to be its matching $y$, then do this
for all elements of first and second parts. On this data set, we could fit the
polynomial equation to the data giving us the $\beta$ values.

\begin{figure}[h]
\center{
  \scalebox{0.55}{
  \includegraphics{./intro/poly.JPG}
  }
}
\caption{Example of Polynomial Fitting}
\vspace{0.6cm}
\end{figure}

For prediction, if we need to estimate $N$ days in the future, then we go back
$N$ days starting from today, hence the historical $N$ points became our new
$x$'s, plugging each $x$ into our ``learned'' polynomial equation would give us
future $y$'s.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Kalman and HMM Markov Models}

In order to obtain better prediction, we hypothesized a combination of HMM and
KF methods could yield different, hopefully better results. We called this
method KMM, a combination of words \textbf{K}F and H\textbf{MM}. There are
various methods in the literature to combine more than one predictor, or
``experts'', for a certain data set. One method is assigning a different expert
for each seperate region, training it, then combining all local experts to
arrive at a better global predictor.

Another method is combining the output of each expert for each data point
prediction. This way, every expert is contributing something to the overall
result, although through different weights. Mathematically every data point
$y_i$ is a combination of
\begin{eqnarray*}
y_i = \sum_j w_j o_{ij}
\end{eqnarray*}
where $o_{ij}$ is prediction from expert $j$ at time $i$, and $w_j$ is the
weight assigned to expert $j$. The sum of weights for all experts must equal to
one, that is, $\sum_j w_j = 1$.

\begin{figure}[h]
\center{
  \scalebox{0.55}{
  \includegraphics{./intro/kmm.jpeg}
  }
}
\caption{KMM Structure}
\vspace{0.6cm}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Previous Work} \label{prevwork}

Using HMM and KF for financial time series prediction is a well known method. As
stated in \cite{ghahramani3} and confirmed by a quick look in to the literature
shows that most probabilistic time series models of use today are descendants of
either HMMs or state space models (SSMs). There is one approach that combines
both HMM and KF methods proposed by Ghahramani and Hinton. They divide the data
into multiple regions detected by HMM, then for each segment, a SSM is trained
as an ``expert'' for that region. In other words, for any given data point, SSMs
are competing among each other to describe that data point. Once SSMs are
trained, the overall flow over those ``multiple regions'' -multiple SSMs- can be
handled by the HMM. There is a similar method called Hidden Markov Experts (HME)
proposed by Shi and Weigend \cite{shi}. HME also utilizes seperation of regions
using HMMs but connects them by using ANNs \cite{nabney}.

Gürgen and Yümlü \cite{gurgen}, employed neural mixture of experts (MOE)
structure for ISE prediction and compared those results with an ANN, a
polynomial and a radial basis function (RBF) nets.

Hassan and Nath \cite{hassan} use HMMs in a way similar that is used in speech
recognition. They decided to train one HMM {\em per trading day} and used four
values of each day in sequence as a mini-time series. Those values were opening
price, high price, low price and closing price. The modeler trains one HMM per
trading day using the four values and also remembers the next day's closing
price. This will become useful for prediction, during which, four values of
yesterday are passed to every trained HMM one by one. Then, whichever HMM
reports the highest likelihood for those four mini time series points, that HMM
is picked as the correct HMM to predict tomorrow. Of course even though the
match gives a change pattern that might be similar to values from yesterday, the
absolute values of both sides can be different - in which case prediction
program simply ``adds'' the average difference of values from matched HMM to
tomorrows predicted stock price. This method \cite{hassan} predicts only one day
ahead.

### 2015-07-26 ms-thesis/ver1/kalman/kalman.tex
-

\documentclass[a4paper,11pt]{article}

\usepackage[latin5]{inputenc}
\usepackage{amsmath}
\usepackage{examplep}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{graphics}

\title{Kalman Filters: Derivation}

\author{Burak Bayramlý}

\begin{document}

\maketitle

Compiling an exhaustive list of fields where Kalman Filters (KF) are in use
would be a formiddable task: The list could start with navigation, radar
tracking, satellite orbit computation \cite{welling}, continue with sonar
ranging, stock price prediction (which hopefully this thesis will prove), and
many other fields where there is a dynamical system present. When the {\em
  Eagle} module of the Apollo 11 mission landed on the moon, it is widely known
that it did so with the help of a Kalman Filter. KF was first proposed in a
seminal paper by \cite{kalman}, later Kalman and Bucy \cite{kalman2} - the
method's first application setting was space tracking.

Kalman Filters, or in general, SSM methods try to model dynamic data using
two-level graphic model. In the first level, there are hidden states, one per
time step, which are random variables distributed as Gaussians. An outside
observer (modeler) does not see these states directly, instead sees them through
an ``observed'' variable whose outputs are assumed to be distorted by another
Gaussian distribution, refered to as white noise. In each time step, it is
assumed the hidden variables evolve linearly, in linear algebra this means the
hidden state is multiplied by a constant matrix $A$. Mathematically,
\begin{eqnarray*}
x_{t+1} &=& Ax_t + v_t 
\end{eqnarray*}
where
\begin{itemize}
   \item $v_0,v_1,...$ is white Gaussian noise with $\mathbf{E}v_t = 0$ and $cov
   \: v_t = Q$
   \item $x_t \sim N(\mu_t, \Sigma_t)$.
\end{itemize}
In order to model the output variables, we multiply hidden states with another
constant matrix and add another Gaussian white noise. All together our SSM looks
as follows:
\begin{eqnarray}
x_{t+1} = Ax_t + v_t \label{m1}\\
y_t = Cx_t + w_t \label{m2}
\end{eqnarray}
where $w_0,w_1,...$ are white Gaussian noise with $\mathbf{E}w_t = 0$ and $cov
\:\: w_t = R$. Random variables $w_t$ and $v_t$ are independent of $x_t$
and $y_t$, and from eachother, this assumption is put forth for simplicity. The
dimensions of each variable are:

\begin{itemize}
   \item $x(t) \in \mathbb{R}^n$ is the state
   \item $y(t) \in \mathbb{R}^p$ is the observed output
   \item $v(t) \in \mathbb{R}^n$ is process noise
   \item $w(t) \in \mathbb{R}^p$ is measurement noise
\end{itemize}

Kalman Filters carry the same conditional independence properties that
characterized HMMs - given a state for time $t$, the future is conditionally
independent from the past, hence KF also carries the first-order Markovian
property just like HMMs. Structurally there are many similarities between two
models as well and its methods for inference, as we shall soon see, the main
difference being that HMM uses multinomial distribution for hidden states
whereas KF uses Gaussians.

Inference with Kalman Filters means estimating the posterior probability given
an observed output sequence. There are two tasks here, filtering and
smoothing. Filtering is formulized by $p(x_t|y_0,...,y_t)$ which tries to
calculate posterior probability for hidden states using measurements up to time
$t$. In smoothing, we will calculate the same probability using measurements
taken after time $t$, which gives us $p(x_t|y_{t+1},...,y_T)$. We also hope to
formulize recursive formulas for both filtering and smoothing.

Both filtering and smoothing have direct consequences in terms of real-life
applications. KFs' natural roots are embedded in space tracking, the method
started as an ``online'' algorithm trying to track down targets using noisy
measurements, as each new measurement is received. In this sense, KF can be seen
as an estimator trying to guess a ``hidden'' position at time $t$ using
measurements up to time $t$. Smoothing, on the other hand, can be thought of a
global ``corrector'' - once we have better knowledge on hand, through smoothing
we can go back to old estimations and correct previous online estimates using
backward recursion.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Filtering}

Following the notation and the derivation steps used by Bishop and Jordan
\cite{jordan}, let's first denote

\begin{itemize}
   \item $\hat{x}_{t}^t \triangleq \mathbf{E}[x_t|y_0,...,y_t]$
   \item $P_{t}^t \triangleq \mathbf{E}[(x_t - \hat{x}_{t|t})(x_t -
   \hat{x}_{t|t})'|y_0,...,y_t]$ 
\end{itemize}
Our first goal is going to be trying to calculate $\hat{x}_{t+1}^{t+1}$ and
$P_{t+1}^{t+1}$ using a new measurement $y_{t+1}$. For this, we need to
``reverse the arrow'' so to speak, in other words formulize the distribution
where $x_t$ has conditional dependence on $y_t$ instead of other way around,
which is what the base equations state right now.

If we take expectation of both sides of \eqref{m1}
\begin{eqnarray*}
\mathbf{E}x_{t+1} = \hat{x}_{t+1} = A\mu_t = A\hat{x}_t
\end{eqnarray*}
Taking covariance of both sides of \eqref{m1} and denoting $P_t$ as $cov \: x(t)$
\begin{eqnarray*}
P_{t+1} = AP_{t}A' + Q
\end{eqnarray*}
This transition is called a ``time update''. It will allow us to propagate the
normal distribution at time $t$ to time $t_1$. The formulas that contain given
$y$'s use the same technique;
\begin{eqnarray*}
\hat{x}_{t+1}^t = Ax_{t}^t\\
P_{t+1}^t = AP_{t}^tA' + Q
\end{eqnarray*}
\begin{eqnarray*}
y_{t+1} &=& Cx_{t+1} + w_t\\
\mathbf{E}[y_{t+1}|y_0,...,y_t] &=& \mathbf{E}[Cx_{t+1}+w_t|y_0,...,y_t]\\
\hat{y}_{t+1}^t &=& C\hat{x}_{t+1}
\end{eqnarray*}
Similarly for covariance
\begin{eqnarray*}
E[(y_{t+1}-\hat{y}_{t+1}^t)(y_{t+1}-\hat{y}_{t+1}^t)'|y_0,...,y_t] =
CP_{t+1}^tC' + R
\end{eqnarray*}

Now the harder task of reversing the arrow: If our goal is formulizing
$p(x_t|y_t)$, then we need to derive the joint distribution between both
variables. Addition of Gaussians is another Gaussian, we can surmize the joint
probability $p(x_t,y_t)$ will be one huge Gaussian.

In order to define the joint normal distribution of $x_{t}$ and $y_{t}$, we need
to write the joint mean and covariance of this big Gaussian. Writing down the
mean will be easy, covariance will be little harder to derive. We can use a
trick \cite{lall} where we write $y_{t} = Cx_{t} + w_t$ as $z = Hu$
where
\begin{eqnarray*}
z =   \left[ \begin{array}{c}
      x_{t} \\
      y_{t}
      \end{array} \right],
H =   \left[ \begin{array}{cc}
      I & 0 \\
      C & I
      \end{array} \right]
u =   \left[ \begin{array}{c}
      x_{t} \\
      w_t
      \end{array} \right],      
\end{eqnarray*}
Now we need to take the covariance of the simpler equation:
\begin{eqnarray*}
cov(z) &=& H cov(u) H'\\
cov(u) &=& \left[ \begin{array}{cc}
      P_{t} & 0 \\
      0 & R
      \end{array} \right]     
\end{eqnarray*}
The full product is:
\begin{eqnarray*}
\left[ \begin{array}{cc}
      I & 0 \\
      C & I
      \end{array} \right]
\left[ \begin{array}{cc}
      P_{t} & 0 \\
      0 & R
      \end{array} \right]
\left[ \begin{array}{cc}
      I & C' \\
      0 & I
      \end{array} \right]           
\end{eqnarray*}
which results in
\begin{eqnarray*}
\left[ \begin{array}{cc}
      P_{t} & P_{t}C' \\
      CP_{t} & CP_{t}C'+R
      \end{array} \right]           
\end{eqnarray*}
We can also re-write this for conditional and with the mean;
\begin{eqnarray}
\left[ \begin{array}{c}
      \hat{x}_{t}^t \\
      C\hat{x}_{t}^t
      \end{array} \right]
\:and\:      
\left[ \begin{array}{cc}
      P_{t}^t & P_{t}^tC' \\
      CP_{t}^t & CP_{t}^tC'+R
      \end{array} \right] \label{jointxy}
\end{eqnarray}
Same for the joint distribution for $x_{t+1}, y_{t+1}$.
\begin{eqnarray}
\left[ \begin{array}{c}
      \hat{x}_{t+1}^t \\
      C\hat{x}_{t+1}^t
      \end{array} \right]
\:and\:      
\left[ \begin{array}{cc}
      P_{t+1}^t & P_{t+1}^tC' \\
      CP_{t+1}^t & CP_{t+1}^tC'+R
      \end{array} \right] \label{jointxy2}
\end{eqnarray}

Now in order to get statements for mean and variance for $x_{t+1}^{t+1}$, we
need to understand partitioned Gaussians. This will guide us when we use
portions of \eqref{jointxy2} and deriving the final equation \cite{jordan}.

An $n$ dimensional Gaussian distribution can be partitioned into $p$ and $q$
dimensional sub-distributions where $n = p + q$. Hence we can say,

\begin{eqnarray}
\mu = 
\left[ \begin{array}{c}
      \mu_1 \\
      \mu_2
      \end{array} \right]           
\Sigma = 
\left[ \begin{array}{cc}
      \Sigma_{11} & \Sigma_{12}\\
      \Sigma_{21} & \Sigma_{22}
      \end{array} \right] \label{condgeneric}
\end{eqnarray}
\begin{eqnarray*}
p(x|\mu,\Sigma) &=& \frac{1}{(2\pi)^{(p+q)/2}|\Sigma|^{1/2}} \\
&& exp\bigg\{
-\frac{1}{2}
\left( \begin{array}{c}
      x_1 - \mu_1 \\
      x_2 - \mu_2
      \end{array} \right)'
\bigg\}
\left[ \begin{array}{cc}
      \Sigma_{11} & \Sigma_{12}\\
      \Sigma_{21} & \Sigma_{22}
      \end{array} \right]^{-1}
\left( \begin{array}{c}
      x_1 - \mu_1 \\
      x_2 - \mu_2
      \end{array} \right)
\bigg\} 
\end{eqnarray*}
After much algebra, we can obtain equation for $p(x_1|x_2)$. From this, we can
get {\em conditioned} $\mu$ and $\Sigma$, as
\begin{eqnarray}
\mu_{1|2} &=& \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}(x_2 - \mu_2) \label{condx} \\
\Sigma_{1|2} &=& \Sigma_{11}-\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21} \nonumber
\end{eqnarray}
Now plugging elements from  \eqref{condx} into \eqref{jointxy2} according
to placement in \eqref{condgeneric} we can formulize $\hat{x}_{t+1}^{t+1}$ and
$P_{t+1}^{t+1}$. 
\begin{eqnarray*}
\hat{x}_{t+1}^{t+1} &=& \hat{x}_{t+1}^t + P_{t+1}^tC'
                  (CP_{t+1}^tC'+ \Sigma_{w})^{-1} \\
               &&   (y_{t+1} - C\hat{x}_{t+1}^t) \\
P_{t+1}^{t+1} &=& P_{t+1}^t - P_{t+1}^tC' (CP_{t+1}^tC' + R)^{-1}CP_{t+1}^t
\end{eqnarray*}
If we declare $K_t \triangleq P_{t+1}^tC' (CP_{t+1}^tC'+ \Sigma_{w})^{-1}$ then,
\begin{eqnarray}
\hat{x}_{t+1}^{t+1} &=& \hat{x}_{t+1}^t + K_t (y_{t+1} -
C\hat{x}_{t+1}^t) \label{xtplus1} \\
P_{t+1}^{t+1} &=& P_{t+1}^t - K_t C P_{t+1}^t
\end{eqnarray}

\begin{thebibliography}{99}

\bibitem{rabiner} Rabiner L. R. , 
  {\em A Tutorial on Hidden Markov Models and Selected Applications in Speech Recognition}, 
  Proceedings of IEEE vol. 77, no. 2, pp. 257-286, 
  1989.

\bibitem{ghahramani} Roweis S. and Z. Ghahramani,
  {\em  A Unifying Review of Linear Gaussian Models },
   Neural Computation 11(2):305--345, 
   1999.

\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
  {\em Parameter Estimation for Linear Dynamical Systems },
   Technical Report CRG-TR-96-2  
   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
   Department of Computer Science, University of Toronto, 
   1996.

\bibitem{ghahramani3} Ghahramani Z., H. E. Hinton,
  {\em Switching State Space Models},
  Technical Report CRG-TR-96-3, Dept. Comp. Sci., Univ. Toronto, 
  1996.

\bibitem{shumway} Shumway R., H. S. Stoffer
  {\em Time series analysis and its applications 2nd Edition},
   New York, Springer, (Springer texts in statistics),
   2000.

\bibitem{jordan} Jordan M. I. , C. Bishop
  {\em An Introduction to Graphical Models},
   Not yet published, 
   2000.

\bibitem{kalman} Kalman R. E.,
  {\em A New Approach to Linear Filtering and Prediction Problems},
  Transactions of the ASME-Journal of Basic Engineering, 82 (Series D): 35-45, 
  1960.

\bibitem{kalman2} Kalman, R.E. and R.S. Bucy, 
  {\em New results in filtering and prediction theory},
  Trans. ASME J. Basic Eng., 83, 95-108,
  1961.

\bibitem{welling} Welling, M.,
  {\em The Kalman Filter - Lecture Tutorial},
   California Institute of Technology,
   2008.

\bibitem{lall} Lall, S.,
  {\em Modern Control 2 Lecture Notes},
   Stanford University,
   2006.

\bibitem{tsay} Tsay, R. S.,
  {\em Time Series and Forecasting: Brief History and Future Research,},
  Journal of the American Statistical Association, Vol. 95, No. 450., pp. 638-643.,
  Jun 2000.

\bibitem{shi} Shi S. and A. S. Weigend., {\em Taking Time Seriously: Hidden
  Markov Experts Applied to Financial Engineering}. In Proceedings of the
  IEEE/IAFE 1997 Conference on Computational Intelligence for Financial
  Engineering , pages 244-252. IEEE, 
  1997.

\bibitem{mandelbrot} Mandelbrot B. B., 
  {\em The variation of certain speculative prices, Journal of Business},
  XXXVI (1963), pp. 392-417, 
  1963.

\end{thebibliography}

\end{document}

### 2015-07-26 ms-thesis/ver1/kalman/ms-thesis-kalman.tex
-

--- 2015-07-26 ms-thesis/ver1/kalman/kalman.tex
@@
-\documentclass[a4paper,11pt]{article}
-
-\usepackage[latin5]{inputenc}
-\usepackage{amsmath}
-\usepackage{examplep}
-\usepackage{amsthm}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{graphics}
-
-\title{Kalman Filters: Derivation}
-
-\author{Burak Bayramlý}
-
-\begin{document}
-
-\maketitle
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\chapter{KALMAN FILTERS}
 
@@
 whereas KF uses Gaussians.
+
+\begin{figure}[!hbp]
+\center{
+  \scalebox{0.6}{
+  \includegraphics{./kalman/kf.JPG}
+  }
+}
+\caption{Hidden and observed states of a KF}
+\end{figure}
 
@@
 
-\begin{thebibliography}{99}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Smoothing}
 
-\bibitem{rabiner} Rabiner L. R. , 
-  {\em A Tutorial on Hidden Markov Models and Selected Applications in Speech Recognition}, 
-  Proceedings of IEEE vol. 77, no. 2, pp. 257-286, 
-  1989.
+Filtering involves applying a new measurement $y_{t+1}$ to calculate
+$\hat{x}_{t+1}^{t+1}$ and $P_{t+1}^{t+1}$ which is an excellent tool for an
+online algorithm. However for stock price prediction, we will have all data we
+need already at hand. As opposed to a tracking application where data cannot be
+known in advance, we will have our data available, as measurements
+$y_1,...,y_T$, in the form of stock prices, to train a Kalman filter based
+solution, all at once. Hence we need an equation for $x_{t}'$. For this we need
+RTS (Rauch-Tung-Striebel) Smoother.
 
-\bibitem{ghahramani} Roweis S. and Z. Ghahramani,
-  {\em  A Unifying Review of Linear Gaussian Models },
-   Neural Computation 11(2):305--345, 
-   1999.
+Based on \eqref{jointxy}, we can write the covariance of joint distribution
+$(x_t,x_{t+1})$. In \eqref{jointxy} the multiplication was with $C$, here it's
+$A$. Hence,
+\begin{eqnarray*}
+\left[ \begin{array}{cc}
+      P_{t}^t & P_{t}^tA' \\
+      AP_{t}^t & AP_{t}^tA'+R
+      \end{array} \right]
+\end{eqnarray*}
+Combining the previous equations together, we need to reach to a statement for
+$x_{t}'$. We can start with $\mathbf{E}[x_t|x_{t+1},y_0,...,y_t)]$.
+\begin{eqnarray*}
+\mu_{1|2} &=& \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}(x_2 - \mu_2)
+\end{eqnarray*}
+which becomes
+\begin{eqnarray*}
+\mathbf{E}[x_t|x_{t+1},y_0,...,y_t)] = \hat{x}_{t}^t+P_{t}^tA'(P_{t+1}^t)^{-1}(x_{t+1}-\hat{x}_{t+1}^t)
+\end{eqnarray*}
+Define $L_t \triangleq P_{t}^tA'(P_{t+1}^t)^{-1}$. Then,
+\begin{eqnarray*}
+\mathbf{E}[x_t|x_{t+1},y_0,...,y_t)] =
+\hat{x}_{t}^t+L_t(x_{t+1}-\hat{x}_{t+1}^t)
+\end{eqnarray*}
+How about $Var[x_t|x_{t+1},y_0,...,y_t]$? If we remember the joint covariance
+for $(x_t,x_{t+1})$
+\begin{eqnarray*}
+ \left[ \begin{array}{cc} 
+      P_{t}^t & P_{t}^tA' \\
+      AP_{t}^t & AP_{t}^tA'+R
+      \end{array} \right] && \\
+\end{eqnarray*}
+Then
+\begin{eqnarray*}
+\Sigma_{1|2} &=& \Sigma_{11}-\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21}
+\end{eqnarray*}
+becomes
+\begin{eqnarray*}
+Var[x_t|x_{t+1},y_0,...,y_t] &=& P_{t}^t - P_{t}^tA'(P_{t+1}^t)^{-1}AP_{t}^t
+\end{eqnarray*}
+Why?
+\begin{eqnarray*}
+\left[ \begin{array}{cc}
+      P_{t}^t & P_{t}^tA' \\
+      AP_{t}^t & AP_{t}^tA'+R
+      \end{array} \right]
+=      
+\left[ \begin{array}{cc}
+      P_{t}^t & P_{t}^tA' \\
+      AP_{t}^t & P_{t+1}^t
+      \end{array} \right]
+\end{eqnarray*}
+If we try to substitute $L_t \triangleq P_{t}^tA'(P_{t+1}^t)^{-1}$ in to the
+equation:
+\begin{eqnarray*}
+Var[x_t|x_{t+1},y_0,...,y_t] &=& P_{t}^t - P_{t}^t A'(P_{t+1}^t)^{-1}AP_{t}^t \\
+&=& P_{t}^t - P_{t}^tA'(P_{t+1}^t)^{-1} (P_{t+1}^t(P_{t+1}^t)^{-1}) AP_{t}^t\\
+&=& P_{t}^t - (P_{t}^tA'(P_{t+1}^t)^{-1}) P_{t+1}^t ((P_{t+1}^t)^{-1} AP_{t}^t)\\
+&=& P_{t}^t - L_t  P_{t+1}^t L_t'
+\end{eqnarray*}
+Due to conditional independence, when we condition on $x_{t+1}$, we make $x_t$
+independent of future observations $y_{t+1},...,y_T$. Hence we can simply say;
+\begin{eqnarray*}
+E[x_t|x_{t+1},y_0,...,y_T]  &=& E[x_t|x_{t+1},y_0,...,y_t]\\
+&=& \hat{x}_{t}^t +L_t(x_{t+1}-\hat{x}_{t+1}^t) \\
+Var[x_t|x_{t+1},y_0,...,y_T]  &=& Var[x_t|x_{t+1},y_0,...,y_t] \\
+&=& P_{t}^t - L_t  P_{t+1}^t L_t'
+\end{eqnarray*}
+We are almost done. If we can drop the $x_{t+1}$ from LHS of both equations, we
+can have our smoothing equations. Due to the so-called Tower Property:
+\begin{eqnarray*}
+\mathbf{E}[X|Z] &=& \mathbf{E}[\mathbf{E}[X|Y,Z]|Z]\\
+Var[X|Z] &=& Var[\mathbf{E}[X|Y,Z]|Z] + \mathbf{E}[Var[X|Y,Z]|Z]
+\end{eqnarray*}
+we can drop $x_{t+1}$ from LHS.
+\begin{eqnarray}
+\hat{x}_{t}' & \triangleq & \mathbf{E}[x_t|y_0,...,y_T] \nonumber\\
+& = & \mathbf{E}[\mathbf{E}[x_t|x_{t+1},y_0,...,y_T]|y_0,...,y_T] \nonumber\\
+& = & \mathbf{E}[\hat{x}_{t}^t +L_t(x_{t+1}-\hat{x}_{t+1}^t)] \nonumber\\
+& = & \hat{x}_{t}^t +L_t(\hat{x}_{t+1}' -\hat{x}_{t+1}^t) \label{xsmooth}
+\end{eqnarray}
+Using similar algebra, we get
+\begin{eqnarray*}
+P_{t}'  = P_{t}^t + L_t(P_{t+1}'-P_{t+1}^t)L_t'
+\end{eqnarray*}
 
-\bibitem{ghahramani2} Ghahramani Z., H. E. Hinton, 
-  {\em Parameter Estimation for Linear Dynamical Systems },
-   Technical Report CRG-TR-96-2  
-   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
-   Department of Computer Science, University of Toronto, 
-   1996.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Training}
 
-\bibitem{ghahramani3} Ghahramani Z., H. E. Hinton,
-  {\em Switching State Space Models},
-  Technical Report CRG-TR-96-3, Dept. Comp. Sci., Univ. Toronto, 
-  1996.
+In order to train a Kalman Filter based predictor, we need to estimate
+parameters $A,C,Q,R,\pi$. These parameters define a KF.
 
-\bibitem{shumway} Shumway R., H. S. Stoffer
-  {\em Time series analysis and its applications 2nd Edition},
-   New York, Springer, (Springer texts in statistics),
-   2000.
+Making use of Markov property, we write our joint probability as;
 
-\bibitem{jordan} Jordan M. I. , C. Bishop
-  {\em An Introduction to Graphical Models},
-   Not yet published, 
-   2000.
+\begin{eqnarray*}
+p(x_0,..,x_t,y_0,..,y_t) =
+p(x_0)\prod_{t=1}^{T-1}p(x_{t+1}|x_t)\prod_{t=0}^{T-1}p(y_t|x_t)
+\end{eqnarray*}
 
-\bibitem{kalman} Kalman R. E.,
-  {\em A New Approach to Linear Filtering and Prediction Problems},
-  Transactions of the ASME-Journal of Basic Engineering, 82 (Series D): 35-45, 
-  1960.
+\begin{eqnarray*}
+p(y_{t}|x_t) &=& exp
+\bigg\{
+-\frac{1}{2}(y_{t}-Cx_t) ' R^{-1}(y_{t}-Cx_t)
+\bigg\}
+(2\pi)^{-\frac{p}{2}}|R|^{-\frac{1}{2}}\\
+p(x_{t+1}|x_t) &=& exp
+\bigg\{
+-\frac{1}{2}(x_{t+1}-Ax_t) ' Q^{-1}(x_{t+1}-Ax_t)
+\bigg\} \\
+&& (2\pi)^{-\frac{k}{2}}|Q|^{-\frac{1}{2}}\\
+p(x_{1}) &=& exp
+\bigg\{
+-\frac{1}{2}(x_{1}-\pi) ' \Sigma_{\pi}^{-1}(x_{1}-\pi)
+\bigg\}
+(2\pi)^{-\frac{k}{2}}|\Sigma_{\pi}|^{-\frac{1}{2}}
+\end{eqnarray*}
+We take log
+\begin{eqnarray*}
+\log p(x,y) &=& \\
+&& -\sum_{t=0}^{T-1}(-\frac{1}{2}(y_{t}-Cx_t) ' R^{-1}(y_{t}-Cx_t)) - \frac{T}{2}\log|R|\\
+&& -\sum_{t=1}^{T-1}(-\frac{1}{2}(x_{t+1}-Ax_t) ' Q^{-1}(x_{t+1}-Ax_t))\\
+&& - \frac{T-1}{2}\log|Q|\\
+&& -\frac{1}{2}(x_{1}-\pi) ' \Sigma_{\pi}^{-1}(x_{1}-\pi) - \frac{T(p+k)}{2}\log|\Sigma_{\pi}|
+\end{eqnarray*}
 
-\bibitem{kalman2} Kalman, R.E. and R.S. Bucy, 
-  {\em New results in filtering and prediction theory},
-  Trans. ASME J. Basic Eng., 83, 95-108,
-  1961.
+In order to estimate parameters, we take derivative of expected log likelihood,
+set to zero and solve
 
-\bibitem{welling} Welling, M.,
-  {\em The Kalman Filter - Lecture Tutorial},
-   California Institute of Technology,
-   2008.
+\begin{eqnarray*}
+Q &=& \mathbf{E}[\log p(x,y|y)] \\
+\frac{\partial Q}{\partial C} &=& -\sum_{t=1}^{T}R^{-1}y_t \hat{x}_{t}' +
+\sum_{t=1}' R^{-1} C P_t = 0
+\end{eqnarray*}
 
-\bibitem{lall} Lall, S.,
-  {\em Modern Control 2 Lecture Notes},
-   Stanford University,
-   2006.
+\begin{eqnarray*}
+C^{new} &=& \bigg( \sum_{t=1}' y_t \hat{x}_t' \bigg) \bigg( P_{t}' \bigg) ^{-1}
+\end{eqnarray*}
 
-\bibitem{tsay} Tsay, R. S.,
-  {\em Time Series and Forecasting: Brief History and Future Research,},
-  Journal of the American Statistical Association, Vol. 95, No. 450., pp. 638-643.,
-  Jun 2000.
+Using similar technique, we can calculate other constants as;
+\begin{eqnarray*}
+R^{new} &=& \frac{1}{T}\sum_{t=1}^T \big( y_ty_t' - C^{new}\hat{x}_ty_t' \big) \\
+A^{new} &=& \big( \sum_{t=2}^T P_{t,t-1})(\sum_{t=2}^TP_{t-1} \big) ^{-1} \\
+Q^{new} &=& \frac{1}{T-1} \big( \sum_{t=2}^T P_t - A^{new}\sum_{t=2}^TP_{t-1,t} \big)\\
+\pi_1^{new} &=& \hat{x}_1
+\end{eqnarray*}
 
-\bibitem{shi} Shi S. and A. S. Weigend., {\em Taking Time Seriously: Hidden
-  Markov Experts Applied to Financial Engineering}. In Proceedings of the
-  IEEE/IAFE 1997 Conference on Computational Intelligence for Financial
-  Engineering , pages 244-252. IEEE, 
-  1997.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Usage}
 
-\bibitem{mandelbrot} Mandelbrot B. B., 
-  {\em The variation of certain speculative prices, Journal of Business},
-  XXXVI (1963), pp. 392-417, 
-  1963.
+While using Kalman Filters for analysis and prediction purposes, one needs to be
+careful about starting condition of the training algorithm. Because of the
+nature of the training algorithm, the parameters $\mu_i, \sigma_i,A,C,Q,R,P$ are
+all used for likelihood calculation/estimation and are {\em themselves} targets
+for estimation. The ``chicken and egg'' nature of the EM training process
+required us to start somewhere, and that somewhere is picking appropiate
+parameters for the parameters mentioned above.
 
-\end{thebibliography}
+We had to determine said values by trial and error, as part of the process where
+we were trying to determine meta parameters such as number of iterations
+required for MC process, the lowest likelihood value threshold (which was used
+to determine when to stop in the training process - if likelihood difference
+fell below this threshold value, we would conclude that training was complete).
 
-\end{document}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Algorithm}
+
+The algorithm that applies forward-backward algorithm and learning through EM is
+described below \cite{ghahramani}.
+
+\begin{figure}[!hbp]
+\center{
+\begin{tabbing}
+{\tt LDS}\={\tt Inf}\={\tt erence}$(Y,A,C,Q,R,x,P)$\\
+\> {\tt for } $t = 1$ {\tt to } $T$ \% forward recursion \\
+\> \> $x_t^{t-1} = Ax_{t-1}^{t-1}$\\
+\> \> $P_t^{t-1} = AP_{t-1}^{t-1}A' + Q$ {\tt if } $t>1$\\
+\> \> $K_t = P_t^{t-1}C'(CP_t^{t-1}C'+R)^{-1}$\\
+\> \> $x_t^t = x_t^{t-1} + K_t(y_t - Cx_t^{t-1})$\\
+\> \> $P_t^t = P_t^{t-1} - K_tCP_t^{t-1}$\\
+\> {\tt end}\\
+\> {\tt initialize } $\hat{P}_{T,T-1}=(I-K_TC)AP_{T-1}^{T-1}$\\
+\> \> {\tt for } $t=T$ {\tt to } $2$ \% backward recursion \\
+\> \> $J_{t-1} = P_{t-1}^{t-1}A'(P_t^{t-1})^{-1}$\\
+\> \> $\hat{x}_{t-1} = x_{t-1}^{t-1} + J_{t-1}(\hat{x}_{t} - Ax_{t-1}^{t-1})$\\
+\> \> $\hat{P}_{t-1} = P_{t-1}^{t-1} + J_{t-1}(\hat{P}_t - P_t^{t-1})J_{t-1}'$\\
+\> \> $\hat{P}_{t,t-1} = P_t^t J_{t-1}' + J_t(\hat{P}_{t+1,t} - AV_t^t)J_{t-1}'$
+   {\tt if } $t<T$\\
+\> {\tt end}\\
+\> {\tt return }$\hat{x}_t, \hat{P}_t, \hat{P}_{t,t-1}$ {\tt for all } $t$
+\end{tabbing}
+}
+\caption{LDS Inference Algorithm}
+\vspace{0.6cm}
+\end{figure}
+
+\begin{figure}[!hbp]
+\center{
+\begin{tabbing}
+{\tt LDS}\={\tt Learn}\=$(Y,k,$\=$\epsilon)$\\ \> {\tt initialize } $A, C, Q, R,
+x_1^0, P_1^0$\\ \> $\alpha = \sum_t y_ty_t'$\\ \> {\tt while change in
+  likelihood >} $\epsilon$\\ \> \> {\tt LDSInference}$(Y,A,C,Q,R,x_1^0,P_1^0)$
+\% E-step\\ \> \> {\tt initialize } $\delta = 0, \gamma = 0, \sigma = 0$\\ \> \>
+   {\tt for } $t = 1$ {\tt to } $T$\\ \> \> \> $\delta = \delta +
+   y_t\hat{x}_t$\\ \> \> \> $\gamma = \gamma + \hat{x}_t\hat{x}_t' +
+   \hat{P}_t$\\ \> \> \> $\beta = \beta + \hat{x}_t\hat{x}_{t-1}' + P_{t,t-1}$
+       {\tt if } $t>1$\\ \> \> {\tt end}\\ \> \> $\gamma_1 = \gamma -
+       \hat{x}_T\hat{x}_T' - \hat{P}_T$\\ \> \> $\gamma_2 = \gamma -
+       \hat{x}_1\hat{x}_1' - \hat{P}_1$\\ \> \> \% M-step\\ \> \> $C = \delta
+       \gamma^{-1}$\\ \> \> $R = (\gamma - C\delta')/T$\\ \> \> $A =
+       \beta\gamma_1^{-1}$\\ \> \> $Q = (\gamma_2 - A\beta') / (T-1)$\\ \> \>
+       $x_1^0 = \hat{x}_1$\\ \> \> $P_1^0 = \hat{P}_1$\\ \> {\tt end}\\ \> {\tt
+         return } $A, C, Q, R, x_1^0, P_1^0$
+\end{tabbing}
+}
+\caption{LDS Learning Algorithm}
+\vspace{0.6cm}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+\section{Prediction with Kalman Filters}
+
+Generating future values with a KF model is somewhat easier than generating
+random values from an HMM. We know state space model assigns one hidden Gaussian
+distribution per data point, which itself is modeled by another Gaussian. In
+comparison, the number of hidden states of an HMM is specified by the modeler
+and is usually less than the number of data points used for training.
+
+The fact that we have as many hidden/observed random variables as data points
+makes the following more intuitive: After the training phase of the KF which
+includes forward recursion and smoothing, we have estimates for the mean and
+variance for all Gaussian distributions. Another knowledge we have at hand is
+the basic time transition equation for KFs - if we have an estimate at $t$ for
+hidden state, we know how to propagate those estimates to the future at
+$t+1$. Therefore, once training is complete, we can simply fast forward to the
+emission Gaussian at time $T$, and using propagation equation generate a new
+observation Gaussian at time $T+1$ which we use to generate a random number that
+represents ``generatively'' an output at time $T+1$.
+
+One point we have to be careful with is propagating the emission Gaussian, not
+simply the hidden Gaussian. In the literature the basic time transition equation
+is for the latter not the former, hence we need to expand the equation for $y_t$
+by taking time transition into account:
+
+Given,
+\begin{eqnarray}
+x_{t+1} = Ax_t + v_t \\
+y_t = Cx_t + w_t
+\end{eqnarray}
+where $v_0,v_1,...$ is white Gaussian noise with $\mathbf{E}v_t = 0$ and $cov \:
+v_t = Q$ and $w_0,w_1,...$ are also white Gaussian noise with $\mathbf{E}w_t =
+0$ and $cov \: w_t = R$, we can determine $y_{t+1}$ as
+\begin{eqnarray}
+y_{t+1} &=& Cx_{t+1} + w_{t+1}\\
+&=& C(Ax_t + v_t) + w_{t+1} \\
+&=& CAx_t + Cv_t + w_{t+1}
+\end{eqnarray}
+which gives us the distribution of the emission Gaussian $y_{t+1}$. The
+conditional expectation and variance of this distribution given $y_{1,..,t}$ is
+\begin{eqnarray}
+y_{t+1} &=& Cx_{t+1} + w_{t+1}\\
+&=& C(Ax_t + v_t) + w_{t+1} \\
+&=& CAx_t + Cv_t + w_{t+1} \\
+\hat{y}_{t+1|t} &=& CA\hat{x}_{t|t} \\
+cov(y_{t+1}|y_{0,...,t}) &=& CA P_{t|t}A'C' + CQC' + R
+\end{eqnarray}
+We can apply this recursive formula to have our KF generate $M$ many random
+observations starting at $T$ up to $T+N$. The algorithm is
+
+\begin{figure}[!hbp]
+\center{
+\begin{tabbing}
+{\tt ob}\={\tt ser}\={\tt vation}\={\tt Genera}\={\tt tion}$(A,C,R,Q,\mu,\Sigma,T,N)$\\
+\> $mu_x = mu[T-1];$\\
+\> $sigma_x = sigma[T-1];$\\
+\> {\tt for} $i = 1$ {\tt to } $N$\\
+\> \> $mu_y = C A mu_x;$\\
+\> \> $sigma_y = CA sigma_x A'C' + CQC' + R;$\\
+\> \> $o[i] = $ {\tt gaussianNumberGeneration}$(mu_y, sigma_y);$\\
+\> \> $mu_x = A mu_x$\\
+\> \> $sigma_x = A  sigma_x A + Q$\\
+\> {\tt end}\\
+\> {\tt return} $o;$
+\end{tabbing}
+}
+\caption{Generating Observations from an LDS Algorithm}
+\vspace{0.6cm}
+\end{figure}

### 2015-07-26 ms-thesis/ver1/results/ms-thesis-results.tex
-

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\chapter{RESULTS}

For both HMM and KF methods, all training points are taken in one continuous
sequence, and training is performed on a window of size $T$. Once training is
completed, predictions can then be made for the next $N$ days starting right
after the training window finished.

As we have seen, When we trained an HMM or KF over a continuous batch of time
series data, what we actually achieved is compressing that data into a shorter,
more compact form. In a sense the data shrinks and the resulting model gives us
a summary of what the data contains. Once this is completed and let us say we
wanted to re-generate data that was used to train the learner, theoretically
this should be possible - for instance for HMMs, starting with the initial
probabilities $\pi$, one can ``roll the dice'' based on $\pi$ and transition
matrix $A$, we can choose a transition, follow that to the next hidden state,
then we throw dice again which lets us follow a transition probability contained
in $A$, so on. This lets us to run a simulation of the data we used to train our
HMM. The real data could have contained 1000 points, however with a trained HMM,
all we really need is $\pi, A, \eta$. The rest is random number generation using
the HMM. For generating {\em future} values, we modified this method a little
further.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Monte Carlo Simulation}

During our tests, we found that executing HMM or KF forward prediction ``once''
did not give optimal results. The reason for the bad performance was hypothized
to be related to pseudorandom number generation - our method, using HMM or KF
relies on pseudorandom number generation over gaussian and uniform
distributions. Especially when a prediction is performed for more than 5 days,
it was observed that the results would degenerate very rapidly. In order to
offset this issue, we decided that using Monte Carlo (MC) method could alleviate
these problems.

Monte Carlo method involves the use of pseudorandom numbers to generate results,
and each experiment is repeated many times to obtain better accuracy and
approximation. For HMM and KF pseudorandom number generation, we chose to follow
this method, then, in each case, the forward prediction which utilizes
pseudorandom number generation was repeated for over $500$ times. After each run
the predicted values of each iteration were summed and averaged for a final list
of predicted values. The iteration numbers were determined during training phase
where many different parameters were tested for optimal results. We observed
that the results improved considerably by this method.

For the generation of pseudorandom numbers, we employ Java
\textsuperscript{\textregistered} language's internal random number
generator. However, an important rule of MC simulation is that the experiments
need to be replicable - that is, each time the same experiment is executed, the
pseudorandom generation should be exactly the same as the previous run. The
default mode of Java pseudorandom number generation takes the current time as
its seed. This would give us different random number for each run. We had to
modify this by setting the seed ourselves manually. A hex value large enough
such as \PVerb!0x5DAE6EE6DL!  yielded good results.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Results}

For our testing we used ISE-100 Index historical data between dates 30/10/1987
and 29/1/2008, for DJI between dates 02/02/1970 and 02/02/2007. Two thirds of
the data was used for training (to adjust model parameters) then the rest of the
data was used for testing final results.

We picked single Gaussian as our HMM's emission distribution and determined 7 to
be the number of hidden states. All HMM, KF and KMM methods used a single
dimension to model their input data. Important parameters used for all methods
are presented below:

\begin{table}[h]
\caption{Parameters}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|}
\hline 
\textbf{Method} & \textbf{Parameter} & \textbf{Value} \\
\hline 
HMM & \# of hidden states  & 7 \\
\hline
ANN - ISE & \# of inputs & 1700 \\
\hline
ANN - DJI & \# of inputs & 3000 \\
\hline
ANN & \# of output & 1 \\
\hline
ANN & \# of Neurons in Hidden Layer & 30 \\
\hline
Polynomial & Degree & 10 \\
\hline
ANN - ISE and DJI & Size of Training Window & 300 \\
\hline
All (except ANN) - ISE & Size of Training Window & 1700 \\
\hline
All (except ANN) - DJI & Size of Training Window & 3000 \\
\hline
\end{tabular}
\end{table}

Kalman/HMM/KMM testing involved a runner program that was able to divide the
data into appropiate pieces, forming one test run, executing it and collecting
the results. In order to increase the number of test runs, we decided to form a
``shifting window'' in the data, running HMM/Kalman/KMM training and prediction
on this window. The training window was shifted 30 times for each run. The
runner starts with a section of data $W$ wide, trains the model, predicts the
next 1,5 and 10 days and compares the results against actual stock price
data. While comparing predicted vs. actual data, the runner program collects
various metrics. A short description of these metrics can be found below;

\subsection{Metrics}

\textbf{MSE Ror}: This metric's long name is mean square error for rate of
returns. In general we can say in our application we collect three types of
metrics: Prediction error on rate of returns, prediction error on absolute
prices and hit rates for increase and decrease predictions. {\em MSE Ror} falls
in the ``rate of returns'' category.  The formula for mse ror is below; $R_i$ is
real rate of returns, $S_i$ is the real stock price, $X_i$ is estimates for rate
of returns or prices on day $i$ and $N$ is the number of predictions, MSE Ror
is:
\begin{eqnarray}
\sqrt{\frac{\sum_i (R_i - X_i)^2}{N}}. 
\end{eqnarray}
\textbf{MAE Ror}: The long name for this metric is ``mean absolute error for
rate of returns''. It calculates the absolute error between predicted rate of
returns and real values. The formula for it is 
\begin{eqnarray}
\frac{\sum_i (|R_i - X_i|)}{N}
\end{eqnarray}
\textbf{MSE Price} statistic is very similar to MSE Ror except that instead of
rate of returns we are using the stock prices themselves. This metric is
formulized as 
\begin{eqnarray}
\sqrt{\frac{\sum_i (S_i - X_i)^2}{N}}
\end{eqnarray}
\textbf{MAE Price} is similar to MAE Ror except that instead of rate of returns
we are using the stock prices themselves which is formulized as
\begin{eqnarray}
\frac{\sum_i (|S_i - X_i|)}{N}
\end{eqnarray}
\textbf{HR} whose long name is ``hit rate'' calculates the number of times
(days) the model succeeded in making a correct increase or decrease prediction
for stock price data. This metric does not look at actual amount of increase in
price, it simply looks at the rate of returns metric and compares the ``signs''
of predicted and actual values. Since rate of returns inherently reveal whether
stock price increased or decreased, the comparison of signs suffices for hit
rate calculations. {\em HR+} captures the number of times an increase was
predicted correctly, {\em HR-} captures the number of times a decrease was
predicted correctly. {\em HR} is a sum of HR+ and HR-.

\subsection{Plots}

To produce the plots seen in \ref{1day1} and \ref{1day2} of ``one day ahead''
predictions, we used a shifting training window $W$ of size $1700$. HMM, KF and
KMM were trained using data points $t,...,t+W$ and using the trained model, we
generated a prediction for the next day, $t+W+1$. Then the training window is
shifted to the right by one day and the same procedure is repeated. This
procedure was performed for 30 days. The predictions, along with the actual
index price for each day are plotted side by side in the same graph. This graph
was inspired by Hassan and Nath \cite{hassan} (covered in Section
\ref{prevwork}) who used it to display their method's results. The comparison of
our graph and theirs can give a good sense of both methods in terms of
prediction success.

\begin{figure}[!hbp]
\center{
  \scalebox{0.35}{
  \includegraphics{./results/1-day-imkb-1.jpg}
  }
}

\caption{\label{1day1} One Day ISE-100 Predictions Using HMM, KF and ANN}
\vspace{0.6cm}
\end{figure}

\begin{figure}[!hbp]
\center{
  \scalebox{0.35}{
  \includegraphics{./results/1-day-imkb-2.jpg}
  }
}
\caption{\label{1day2}One Day ISE-100 Predictions Using polynomial, random and KMM}
\vspace{0.6cm}
\end{figure}

\begin{figure}[!hbp]
\center{
  \scalebox{0.65}{
  \includegraphics{./results/hasannath.JPG}
  }
}
\caption{\label{1day5}One Day Predictions Using HMM method by Hassan \& Nath \cite{hassan}}
\label{1dayhassan}
\vspace{0.6cm}
\end{figure}

\subsection{Conclusion}

\begin{table}[h]
\caption{\label{table1}ISE Prediction - 1 day}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
 &                  HMM &       Kalman &   Random &        ANN &     Polynomial &    KMM\\
\hline
   MSE Ror &      0.011 &      0.011 &      0.049 & 0.01154460 &      0.015 &      0.012\\
\hline
 MSE Price &    114.715 &     36.446 &   2176.878 & 1.24473817 &    518.535 &    156.740\\
\hline
   MAE ROR &      0.011 &      0.011 &      0.049 & 0.01154460 &      0.020 &      0.012\\
\hline
 MAE Price &    114.715 &     36.446 &   2176.878 & 1.24473817 &    518.535 &    156.740\\
\hline
        HR &      0.567 &      0.700 &      0.467 & 0.54838710 &      0.581 &      0.600\\
\hline
        H+ &      0.333 &      0.467 &      0.267 & 0.32258065 &      0.581 &      0.400\\
\hline
        H- &      0.233 &      0.233 &      0.200 & 0.22580645 &      0.000 &      0.200\\
\hline
\end{tabular}
\end{table}

As ``one day ahead'' predictions (seen in Table \ref{table1}) show , HMM, KF and
KMM methods offer results that are better than random and polynomial based
predictions. Kalman filters offer the best prediction in hit rate categories,
but the differences are especially obvious in absolute price categories. The
only contender here was the ANN method, it was able to surpass KF prediction in
terms of {\em MAE} and {\em MSE} metrics. Also, KF predictions in these
categories are almost one third of HMMs. Most promising result was the {\em HR}
result, for KF at 70\%. ANN and polynomial {\rm HR} results were better than
HMM. The winner here are both ANN and KF methods. For trading purposes, I would
use KF because of its success at hit rate metrics.

\begin{table}[h]
\caption{\label{table2}ISE Prediction - 5 days}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
 &                  HMM &       Kalman &   Random &        ANN &     Polynomial &    KMM\\
\hline
   MSE Ror &      0.007 &      0.007 &      0.027 & 0.00656345 &      0.007 &      0.007\\
\hline
 MSE Price &    382.314 &    375.512 &   1727.495 & 273.23645268 &    428.181 &    414.425\\
\hline
   MAE ROR &      0.012 &      0.012 &      0.053 & 0.01199135 &      0.015 &      0.013\\
\hline
 MAE Price &    690.210 &    690.491 &   3415.059 & 444.12065452 &    827.508 &    765.393\\
\hline
        HR &      0.527 &      0.493 &      0.440 & 0.58709677 &      0.490 &      0.467\\
\hline
        H+ &      0.273 &      0.240 &      0.240 & 0.28387097 &      0.490 &      0.213\\
\hline
        H- &      0.253 &      0.253 &      0.200 & 0.30322581 &      0.000 &      0.253\\
\hline
\end{tabular}
\end{table}

As the number of days being predicted increases (seen in Table \ref{table2}), we
notice two things: The error rates in terms of absolute prices start increasing
and second, in terms of {\em HR} metric, the difference between HMM and KF and
their closest contender polynomial method decreases as well. In fact, all
methods start approaching random prediction which in the long run must approach
\%50 success rate. In terms of {\em MAE} metric, ANN method was the top
performer, closely follwed by HMM, KF and KMM methods. One surprise here was
ANN's {\em HR} metric, it was at \%58 which for 5 day prediction is still a very
good result. However we did not witness the same level of success for DJI 5-day
prediction, hence we contributed this success to being ``lucky'', perhaps due to
random number generation outputting just the right values, or to precision
somehow working in the method's favor. The top performer in terms of {\em MSE}
results is again ANN at around \%0.006, so the method's overall performance is
still very good. The difference between the top performers and random prediction
is almost tenfold. Random prediction is doing very badly, which, of course it
should.

\begin{table}[h]
\caption{\label{table3}ISE Prediction - 10 days}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
 &                  HMM &       Kalman &   Random &        ANN &     Polynomial &    KMM\\
\hline
   MSE Ror &      0.005 &      0.005 &      0.019 & 0.00514647 &      0.005 &      0.005\\
\hline
 MSE Price &    407.231 &    424.893 &   1514.800 & 229.06783797 &    456.783 &    437.967\\
\hline
   MAE ROR &      0.013 &      0.013 &      0.053 & 0.01301490 &      0.014 &      0.014\\
\hline
 MAE Price &   1064.323 &   1107.195 &   4116.571 & 547.09327032 &   1239.209 &   1146.643\\
\hline
        HR &      0.500 &      0.497 &      0.437 & 0.52580645 &      0.510 &      0.483\\
\hline
        H+ &      0.270 &      0.240 &      0.223 & 0.25806452 &      0.506 &      0.230\\
\hline
        H- &      0.230 &      0.257 &      0.213 & 0.26774194 &      0.003 &      0.253\\
\hline
\end{tabular}
\end{table}

For 10 day predictions (seen in Table \ref{table3}), we see that all methods
with the exception of random prediction started approaching the same performance
numbers in terms of {\em HR}. This is an indication that prediction 10 days
ahead is a very hard task, and also, that each method is losing its
advantage. Here, again, polynomial method was consistent in its success, however
it also has a problem that its {\em H-} metric which is at a nearly zero. In
comparison its {\em H+} result is at ~\%50, this means the method has a
bias toward increases. This bias would cause problems if one wanted to use this
method for any form of trading activity.

The closeness of the results can be witnessed not only for {\em HR} metric, but
also for absolute price and {\em HR} statistics. ANN method reported the best
prediction in terms of absolute stock prices.

It is clear looking at all {\em HR} numbers that for 10 days, we are at a level
where random prediction is as good as any other method. Since the probability of
guessing an increase and decrease are both ~$0.25$ (in total $0.5$) in the long
run, we can see all methods are in the neighborhood of those numbers, in other
words, predicting the increase and decrease 10 days ahead is as good as flipping
a coin.

\begin{table}[h]
\caption{\label{table4}Dow Jones Prediction - 1 day}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
 &                  HMM &       Kalman &   Random &        ANN &     Polynomial &    KMM\\
\hline
   MSE Ror &      0.005 &      0.005 &      0.058 & 0.00580727 &      0.009 &      0.005\\
\hline
 MSE Price &     10.305 &      3.694 &    654.055 & 0.08456436 &     57.121 &     18.566\\
\hline
   MAE ROR &      0.005 &      0.005 &      0.058 & 0.00580727 &      0.011 &      0.005\\
\hline
 MAE Price &     10.305 &      3.694 &    654.055 & 0.08456436 &     57.121 &     18.566\\
\hline
        HR &      0.600 &      0.633 &      0.500 & 0.57142857 &      0.548 &      0.600\\
\hline
        H+ &      0.367 &      0.367 &      0.233 & 0.33333333 &      0.290 &      0.367\\
\hline
        H- &      0.233 &      0.267 &      0.267 & 0.23809524 &      0.258 &      0.233\\
\hline
\end{tabular}
\end{table}

For ``one day ahead'' predictions using DJI values (seen in Table \ref{table4}),
we see similar performances as we saw with ISE-100 predictions. This actually is
a relief because it proves that financial time series data behaves in similar
ways to ISE time series values, and the trends are the same no matter which
stock, index or country the prices are based on. More similarities can be seen
in all metrics - for example {\em MSE Ror} difference between HMM/KF/KMM and
polynomial prediction are still within 40\%, {\em HR} success of KF method is
the best, closely followed by ANN, HMM and KMM and then polynomial
methods. There is almost a tenfold difference between {\em MSE Ror} for KF and
random prediction. In terms of rate of return prediction, HMM/KF/KMM reported
the best results even though in terms of absolute prices, ANN still performed
better than all of the other methods.

\begin{table}[h]
\caption{\label{table5}Dow Jones Prediction - 5 days}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
 &                  HMM &       Kalman &   Random &        ANN &     Polynomial &    KMM\\
\hline
   MSE Ror &      0.003 &      0.003 &      0.025 & 0.00247295 &      0.003 &      0.003\\
\hline
 MSE Price &     45.048 &     43.990 &    403.774 & 21.49305772 &     51.288 &     47.909\\
\hline
   MAE ROR &      0.005 &      0.005 &      0.050 & 0.00449691 &      0.005 &      0.005\\
\hline
 MAE Price &     83.589 &     82.309 &    797.036 & 35.36699410 &    103.298 &     89.661\\
\hline
        HR &      0.460 &      0.553 &      0.473 & 0.51428571 &      0.587 &      0.460\\
\hline
        H+ &      0.320 &      0.347 &      0.307 & 0.30476190 &      0.329 &      0.293\\
\hline
        H- &      0.140 &      0.207 &      0.167 & 0.20952381 &      0.258 &      0.167\\
\hline
\end{tabular}
\end{table}

And just as was the case with ISE predictions, HMM/KF and polynomial results
start approaching eachother as number of predicted days start to increase (seen
in Table \ref{table5}). However, in terms of DJI prediction, this approach
happen much earlier (at 5 days) than was the case with ISE predictions. With ISE
prediction, it was not until 10 days that the results came close to each other.

\begin{table}[h]
\caption{\label{table6}Dow Jones Prediction - 10 days}
\vspace{0.3cm}
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
 &                  HMM &       Kalman &   Random &        ANN &     Polynomial &    KMM\\
\hline
   MSE Ror &      0.002 &      0.002 &      0.018 & 0.00187652 &      0.002 &      0.002\\
\hline
 MSE Price &     41.802 &     43.171 &    376.496 & 18.27777881 &     44.407 &     41.473\\
\hline
h   MAE ROR &      0.005 &      0.005 &      0.051 & 0.00477930 &      0.005 &      0.005\\
\hline
 MAE Price &    111.533 &    117.051 &   1012.789 & 44.48823690 &    123.517 &    110.753\\
\hline
        HR &      0.510 &      0.557 &      0.537 & 0.47619048 &      0.484 &      0.527\\
\hline
        H+ &      0.320 &      0.283 &      0.287 & 0.28095238 &      0.235 &      0.290\\
\hline
        H- &      0.190 &      0.273 &      0.250 & 0.19523810 &      0.248 &      0.237\\
\hline
\end{tabular}
\end{table}

As one can see from the data tables \ref{table1},\ref{table2},\ref{table3},
\ref{table4},\ref{table5} and \ref{table6} all methods were able to consistenly
outperform random number based prediction. We expected to see those results. In
respect to {\em MSE} and {\em MAE} metrics, ANN, HMM, KF and KMM methods are
found to be superior to other methods. For ``one day ahead'' prediction, KF
based prediction was best in terms of {\em HR} metric. The downside with
polynomial fitting was its {\em H-} metric always reported values close to zero,
which means this method was not able to guess market decreases meaning it is
biased. HMM, KF, KMM and ANN methods all have a healthy allocation between {\em
  H+} and {\em H-} values.

The ``one day ahead'' plots make the success of the guesses quite visible as
well. One point that needs mentioning here is that it could be misleading to
compare our graph directly with Hassan and Nath's, because they have
consistently used the four values mentioned in order to ``find'' an HMM whereas
we used one single HMM to ``generate'' a rate of return which we use to multiply
with the last days' ``known'' price.

In terms of trading, various strategies can be devised in order to maximize
return on investment. A simple trading strategy that takes a single stock or
index as a basis for trade could decide to ``buy'' a stock if it predicts
increase, or a``sell'' if it predicts a decrease for the next day. In case of
``buy'' and ``sell'' sequence, the difference would be profit. In case of
``sell'' and ``buy'' sequence. In other words we could still make money, by
extending this strategy in order to make money from ``decreases'' in the
market. Most markets offer a trading activity called ``short selling'' which
allows a trader to sell a security he does not own, in hope of repurchasing them
at a lower price. If I know the price of a security decrease to \$100 tomorrow,
and the price is \$120 today, I can shortsell at \$120, with a commitment to buy
the actual stock tomorrow. Then if the price falls, I will be buying at a lower
price selling it higher, pocketing the difference as profit.

Overall we can conclude HMM/KF/KMM methods performed reasonably well and they
can be used as a basis for a trading system.

### 2015-07-26 ms-thesis/ver2/biblio/bib.tex
-

--- 2015-07-26 ms-thesis/ver1/biblio/bib.tex
@@
    Technical Report CRG-TR-96-2  
-   \PVerb![ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz]!,
+   \url{ftp://ftp.cs.toronto.edu/pub/zoubin/tr96-2.ps.gz},
    Department of Computer Science, University of Toronto, 
@@
 
-\bibitem{alpaydin} Alpaydın E., 
+\bibitem{alpaydin} Alpaydin E., 
   {\em Introduction to Machine Learning},
@@
   {\em Modelling Financial Time Series with Switching State Space Models},
-  \PVerb!citeseer.ist.psu.edu/267552.html!,
+  \url{citeseer.ist.psu.edu/267552.html},
   1999.

### 2015-07-26 ms-thesis/ver2/cover/ms-thesis-cover.tex
-

--- 2015-07-26 ms-thesis/ver1/cover/ms-thesis-cover.tex
@@
-\title{FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN MARKOV MODELS}
-\turkcebaslik{HMM VE KALMAN FİLTRELERİ KULLANARAK FİNANSAL ZAMAN SERİSİ TAHMİNİ}
+\title{FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN
+  MARKOV MODELS \\ \\ HMM VE KALMAN FILTRELERI KULLANARAK FINANSAL ZAMAN SERISI TAHMINI}
 
@@
 
-\author{Burak Bayramlı}
+\author{Burak Bayramli}
 
@@
 
-\supervisor{Prof. Fikret Gürgen}
+\supervisor{Prof. Fikret Gurgen}
 
-\examineri{Prof. Levent Akın}
+\examineri{Prof. Levent Akin}
 
-\examinerii{Prof. Ethem Alpaydın}
+\examinerii{Prof. Ethem Alpaydin}
 

### 2015-07-26 ms-thesis/ver2/hmm/ms-thesis-hmm.tex
-

--- 2015-07-26 ms-thesis/ver1/hmm/ms-thesis-hmm.tex
@@
 probabilities at discrete intervals. For the simplest, first-order Markov Chain,
-these probabilities can be captured in a \PVerb!NxN!  matrix. The fact that
+these probabilities can be captured in a \verb!NxN!  matrix. The fact that
 present state is enough to represent a system and is the only factor that
@@
 A Markov Chain could also be second order or third order and so on - for example
-for second order Markov Chains, the transition table is \PVerb!2N x N! which
+for second order Markov Chains, the transition table is \verb!2N x N! which
 necessitates both present and {\em previous} step to obtain a transition probability
@@
   \scalebox{0.8}{
-  \includegraphics{./hmm/hmm-intro.jpg}
+  \includegraphics{./hmm/hmm-intro.JPG}
   }
@@
   \scalebox{0.8}{
-  \includegraphics{./hmm/hmm-dep.jpg}
+  \includegraphics{./hmm/hmm-dep.JPG}
   }
@@
 \begin{eqnarray*}
-\alpha(q_t) & \triangleq & p(y_0...y_t,q_t) \\
-\beta(q_t) & \triangleq & p(y_{t+1}...y_T|q_t)
+\alpha(q_t) & \equiv & p(y_0...y_t,q_t) \\
+\beta(q_t) & \equiv & p(y_{t+1}...y_T|q_t)
 \end{eqnarray*}
@@
 the transition table. The pseucode for the function
-\PVerb!observationGeneration!  looks as follows:
+\verb!observationGeneration!  looks as follows:
 
@@
 \> \> {\tt if} $(randVal > total$ {\tt and } \\
-\> \> \> $randVal \leqslant (total + A[currentState][j]))$\\
+\> \> \> $randVal \leq (total + A[currentState][j]))$\\
 \> \> \> $currentState = j;$\\

### 2015-07-26 ms-thesis/ver2/hmm/proof2.tex
-

There is a little complication for the start-up formula for $\beta$
recursion. In this case it would not be possible to start with $\beta(q_{T+1})$
because that state does not exist in our system. In that case, could we start
with $\beta(q_T)$ equaling vector of ones?

\begin{proof}
From \eqref{py} we can use the formula for $p(y)$ at time $T$ to confirm this.

\begin{eqnarray*}
p(y) &=& \sum_{q_T} \alpha(q_T)\beta(q_T) \\
&=& \sum_{i} \alpha(q_T^i=1)\beta(q_T^i=1) 
\end{eqnarray*}

Since we assumed $\beta(q_T)$ would be vector of ones, $\beta(q_T^i=1)$ will
always be 1, hence it can be dropped

\begin{eqnarray*}
&=& \sum_{i} \alpha(q_T^i=1)\\
&=& \sum_{i} p(y_0,...,y_T,q_T^i=1)\\
&=& p(y_0,...,y_T)\\
&=& p(y)
\end{eqnarray*}

which verifies our assumption.
\end{proof}

### 2015-07-26 ms-thesis/ver2/kalman/ms-thesis-kalman.tex
-

--- 2015-07-26 ms-thesis/ver1/kalman/ms-thesis-kalman.tex
@@
 \begin{eqnarray}
-x_{t+1} = Ax_t + v_t \label{m1}\\
-y_t = Cx_t + w_t \label{m2}
+x_{t+1} = Ax_t + v_t \mlabel{1}\\
+y_t = Cx_t + w_t \mlabel{2}
 \end{eqnarray}
@@
 whereas KF uses Gaussians.
-
-\begin{figure}[!hbp]
-\center{
-  \scalebox{0.6}{
-  \includegraphics{./kalman/kf.JPG}
-  }
-}
-\caption{Hidden and observed states of a KF}
-\end{figure}
 
@@
 \begin{itemize}
-   \item $\hat{x}_{t}^t \triangleq \mathbf{E}[x_t|y_0,...,y_t]$
-   \item $P_{t}^t \triangleq \mathbf{E}[(x_t - \hat{x}_{t|t})(x_t -
+   \item $\hat{x}_{t}^t \equiv \mathbf{E}[x_t|y_0,...,y_t]$
+   \item $P_{t}^t \equiv \mathbf{E}[(x_t - \hat{x}_{t|t})(x_t -
    \hat{x}_{t|t})'|y_0,...,y_t]$ 
@@
 
-If we take expectation of both sides of \eqref{m1}
+If we take expectation of both sides of (1)
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-Taking covariance of both sides of \eqref{m1} and denoting $P_t$ as $cov \: x(t)$
+Taking covariance of both sides of (1) and denoting $P_t$ as $cov \: x(t)$
 \begin{eqnarray*}
@@
       CP_{t}^t & CP_{t}^tC'+R
-      \end{array} \right] \label{jointxy}
+      \end{array} \right] \mlabel{3}
 \end{eqnarray}
@@
       CP_{t+1}^t & CP_{t+1}^tC'+R
-      \end{array} \right] \label{jointxy2}
+      \end{array} \right] \mlabel{4}
 \end{eqnarray}
@@
 need to understand partitioned Gaussians. This will guide us when we use
-portions of \eqref{jointxy2} and deriving the final equation \cite{jordan}.
+portions of (4) and deriving the final equation \cite{jordan}.
 
@@
       \Sigma_{21} & \Sigma_{22}
-      \end{array} \right] \label{condgeneric}
+      \end{array} \right] \mlabel{6}
 \end{eqnarray}
@@
 \begin{eqnarray}
-\mu_{1|2} &=& \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}(x_2 - \mu_2) \label{condx} \\
+\mu_{1|2} &=& \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}(x_2 - \mu_2) \mlabel{5} \\
 \Sigma_{1|2} &=& \Sigma_{11}-\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21} \nonumber
 \end{eqnarray}
-Now plugging elements from  \eqref{condx} into \eqref{jointxy2} according
-to placement in \eqref{condgeneric} we can formulize $\hat{x}_{t+1}^{t+1}$ and
+Now plugging elements from  (5) into (4) according
+to placement in (6) we can formulize $\hat{x}_{t+1}^{t+1}$ and
 $P_{t+1}^{t+1}$. 
@@
 \end{eqnarray*}
-If we declare $K_t \triangleq P_{t+1}^tC' (CP_{t+1}^tC'+ \Sigma_{w})^{-1}$ then,
+If we declare $K_t \equiv P_{t+1}^tC' (CP_{t+1}^tC'+ \Sigma_{w})^{-1}$ then,
 \begin{eqnarray}
@@
 
-Based on \eqref{jointxy}, we can write the covariance of joint distribution
-$(x_t,x_{t+1})$. In \eqref{jointxy} the multiplication was with $C$, here it's
+Based on (3), we can write the covariance of joint distribution
+$(x_t,x_{t+1})$. In (3) the multiplication was with $C$, here it's
 $A$. Hence,
@@
 \end{eqnarray*}
-Define $L_t \triangleq P_{t}^tA'(P_{t+1}^t)^{-1}$. Then,
+Define $L_t \equiv P_{t}^tA'(P_{t+1}^t)^{-1}$. Then,
 \begin{eqnarray*}
@@
 \end{eqnarray*}
-If we try to substitute $L_t \triangleq P_{t}^tA'(P_{t+1}^t)^{-1}$ in to the
+If we try to substitute $L_t \equiv P_{t}^tA'(P_{t+1}^t)^{-1}$ in to the
 equation:
@@
 \begin{eqnarray}
-\hat{x}_{t}' & \triangleq & \mathbf{E}[x_t|y_0,...,y_T] \nonumber\\
+\hat{x}_{t}' & \equiv & \mathbf{E}[x_t|y_0,...,y_T] \nonumber\\
 & = & \mathbf{E}[\mathbf{E}[x_t|x_{t+1},y_0,...,y_T]|y_0,...,y_T] \nonumber\\

### 2015-07-26 ms-thesis/ver2/build.py

@@ -1,0 +1,7 @@
+import os, sys
+
+if len(sys.argv) == 1 or sys.argv[1] == 'tex':
+    os.system("pdflatex -shell-escape ms-the*.tex")
+    os.system("evince ms-the*.pdf")
+    exit()
+    

### 2015-07-26 ms-thesis/ver2/intro/ms-thesis-intro.tex
-

--- 2015-07-26 ms-thesis/ver1/intro/ms-thesis-intro.tex
@@
 
-ISE and DJI data sets consisted of daily closing prices for over 21 years for
-ISE and 30 years for DJI. DJI data was obtained from
-\PVerb!http://finance.yahoo.com! site. ISE-100 historical data is available from
-Turkish Central Bank site at \PVerb!http://evds.tcmb.gov.tr!. We attempted to
-predict future price for 1, 5, and 10 days for each method mentioned above. All
-methods were given a limited ``time window'' on which training took place, after
-which, prediction for future prices could be performed. We kept the training
-window the same size for all algorithms with the exception of ANN because too
-much data was causing overfitting and effecting results negatively.
+ISE and DJI data sets consisted of daily closing prices for over 21 years
+for ISE and 30 years for DJI. DJI data was obtained from
+\url{http://finance.yahoo.com} site. ISE-100 historical data is available
+from Turkish Central Bank site at \url{http://evds.tcmb.gov.tr}. We
+attempted to predict future price for 1, 5, and 10 days for each method
+mentioned above. All methods were given a limited ``time window'' on which
+training took place, after which, prediction for future prices could be
+performed. We kept the training window the same size for all algorithms
+with the exception of ANN because too much data was causing overfitting and
+effecting results negatively.
 
@@
 \end{figure}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+import pandas as pd
+df = pd.read_csv('../data/ise.csv',index_col='Date')
+df['Adj Close'].head(1000).plot()
+plt.hold(False)
+plt.savefig('ms-thesis-intro_01.png')
+plt.hold(False)
+\end{minted}
+
+\includegraphics[height=6cm]{ms-thesis-intro_01.png}
 
@@
 which aims to minimize the error $\epsilon$. At the end this process outputs all
-$\beta$ coeffecients to define the equation for $y$. Matlab's \PVerb!polyfit!
+$\beta$ coeffecients to define the equation for $y$. Matlab's \verb!polyfit!
 accomplished this task.

### 2015-07-26 ms-thesis/ver2/data/yahoo.py
-

@@ -1,0 +1,14 @@
+import datetime
+import pandas as pd
+import pandas.io.data
+
+# Yahoo Finance
+df = pd.io.data.get_data_yahoo('XU100.IS', 
+                                 start=datetime.datetime(1999, 4, 29), 
+                                 end=datetime.datetime(2015, 6, 1))
+df.to_csv('ise.csv')
+
+# FRED
+#jpy = pd.io.data.DataReader('DEXUSEU', 'fred')
+#jpy.to_csv('/tmp/DEXJPUS.csv')
+

### 2015-07-26 ms-thesis/ver2/intro/ms-thesis-intro.tex
-

@@
 import pandas as pd
-df = pd.read_csv('../data/ise.csv',index_col='Date')
-df['Adj Close'].head(1000).plot()
+df = pd.read_csv('../data/dji.csv',index_col='Date')
+df['Adj Close'].head(100).plot()
+plt.hold(False)
 plt.hold(False)
@@
 \includegraphics[height=6cm]{ms-thesis-intro_01.png}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+import pandas as pd
+df = pd.read_csv('../data/ise.csv',index_col='Date')
+df2 = df[df.index <'22-06-2000']
+df2['Adj Close'].plot()
+plt.hold(False)
+plt.savefig('ms-thesis-intro_02.png')
+plt.hold(False)
+\end{minted}
+
+\includegraphics[height=6cm]{ms-thesis-intro_02.png}
 

### 2015-07-26 ms-thesis/ver2/data/yahoo.py

@@ -2,13 +2,13 @@ import datetime
 import pandas as pd
 import pandas.io.data
 
-# Yahoo Finance
 df = pd.io.data.get_data_yahoo('XU100.IS', 
                                  start=datetime.datetime(1999, 4, 29), 
-                                 end=datetime.datetime(2015, 6, 1))
+                                 end=datetime.datetime(2015, 7, 26))
 df.to_csv('ise.csv')
 
-# FRED
-#jpy = pd.io.data.DataReader('DEXUSEU', 'fred')
-#jpy.to_csv('/tmp/DEXJPUS.csv')
+df = pd.io.data.get_data_yahoo('^DJI', 
+                                 start=datetime.datetime(1990, 8, 28), 
+                                 end=datetime.datetime(2015, 7, 26))
+df.to_csv('dji.csv')
 

### 2015-07-26 ms-thesis/ver2/intro/ms-thesis-intro.tex
-

--- 2015-07-26 ms-thesis/ver2/intro/ms-thesis-intro.tex
@@
 
+\begin{minted}[fontsize=\footnotesize]{python}
+import pandas as pd
+df = pd.read_csv('../data/dji.csv',index_col='Date')
+df['Adj Close'].head(100).plot()
+plt.hold(False)
+plt.hold(False)
+plt.savefig(./intro/ms-thesis-intro_01.png)
+plt.hold(False)
+\end{minted}
+
+%\includegraphics[height=6cm]{ms-thesis-intro_01.png}
+
 \begin{figure}[ht]
 \center{
-  \scalebox{0.30}{
-  \includegraphics{./intro/somevals.jpg}
+  \scalebox{0.50}{
+  \includegraphics{./intro/ms-thesis-intro_01.png}
   }
 }
-\caption{Some Values from DJI and ISE-100}
 \vspace{0.6cm}
@@
 df = pd.read_csv('../data/ise.csv',index_col='Date')
-df['Adj Close'].head(1000).plot()
+df2 = df[df.index <'22-06-2000']
+df2['Adj Close'].plot()
 plt.hold(False)
-plt.savefig('ms-thesis-intro_01.png')
+plt.savefig(./intro/ms-thesis-intro_02.png)
 plt.hold(False)
@@
 
-\includegraphics[height=6cm]{ms-thesis-intro_01.png}
+%\includegraphics[height=6cm]{ms-thesis-intro_02.png}
+
+\begin{figure}[ht]
+\center{
+  \scalebox{0.50}{
+  \includegraphics{./intro/ms-thesis-intro_02.png}
+  }
+}
+\vspace{0.6cm}
+\end{figure}
 

### 2015-07-26 ms-thesis/ver2/build.py

@@ -1,7 +1,7 @@
 import os, sys
 
 if len(sys.argv) == 1 or sys.argv[1] == 'tex':
-    os.system("pdflatex -shell-escape ms-the*.tex")
+    os.system("pdflatex -shell-escape ms-thesis.tex")
     os.system("evince ms-the*.pdf")
     exit()
     

### 2015-07-26 ms-thesis/ver2/intro/ms-thesis-intro.tex
-

@@
 import pandas as pd
-df = pd.read_csv('../data/dji.csv',index_col='Date')
-df['Adj Close'].head(100).plot()
+dji = pd.read_csv('../data/dji.csv',index_col='Date')
+dji['Adj Close'].head(100).plot()
 plt.hold(False)
@@
 import pandas as pd
-df = pd.read_csv('../data/ise.csv',index_col='Date')
-df2 = df[df.index <'22-06-2000']
-df2['Adj Close'].plot()
+ise = pd.read_csv('../data/ise.csv',index_col='Date')
+ise2 = ise[df.index <'22-06-2000']
+ise2['Adj Close'].plot()
 plt.hold(False)
-plt.savefig(./intro/ms-thesis-intro_02.png)
+plt.savefig('ms-thesis-intro_02.png')
 plt.hold(False)
@@
 
+\begin{minted}[fontsize=\footnotesize]{python}
+ise2['Returns'] = ise2['Adj Close'].pct_change()
+ise2.Returns.hist()
+ise2.Returns.plot()
+plt.hold(False)
+plt.savefig('ms-thesis-intro_03.png')
+plt.hold(False)
+\end{minted}
+
+%\includegraphics[height=6cm]{ms-thesis-intro_03.png}
+
 \begin{figure}[!hbp]
 \center{
-  \scalebox{0.30}{
-  \includegraphics{./intro/volimkb.jpg}
+  \scalebox{0.70}{
+  \includegraphics{./intro/ms-thesis-intro_03.png}
   }

### 2015-07-27 ms-thesis/ver2/kalman/ms-thesis-kalman.tex
-

@@
 
-\begin{figure}[!hbp]
-\center{
-\begin{tabbing}
-{\tt LDS}\={\tt Inf}\={\tt erence}$(Y,A,C,Q,R,x,P)$\\
-\> {\tt for } $t = 1$ {\tt to } $T$ \% forward recursion \\
-\> \> $x_t^{t-1} = Ax_{t-1}^{t-1}$\\
-\> \> $P_t^{t-1} = AP_{t-1}^{t-1}A' + Q$ {\tt if } $t>1$\\
-\> \> $K_t = P_t^{t-1}C'(CP_t^{t-1}C'+R)^{-1}$\\
-\> \> $x_t^t = x_t^{t-1} + K_t(y_t - Cx_t^{t-1})$\\
-\> \> $P_t^t = P_t^{t-1} - K_tCP_t^{t-1}$\\
-\> {\tt end}\\
-\> {\tt initialize } $\hat{P}_{T,T-1}=(I-K_TC)AP_{T-1}^{T-1}$\\
-\> \> {\tt for } $t=T$ {\tt to } $2$ \% backward recursion \\
-\> \> $J_{t-1} = P_{t-1}^{t-1}A'(P_t^{t-1})^{-1}$\\
-\> \> $\hat{x}_{t-1} = x_{t-1}^{t-1} + J_{t-1}(\hat{x}_{t} - Ax_{t-1}^{t-1})$\\
-\> \> $\hat{P}_{t-1} = P_{t-1}^{t-1} + J_{t-1}(\hat{P}_t - P_t^{t-1})J_{t-1}'$\\
-\> \> $\hat{P}_{t,t-1} = P_t^t J_{t-1}' + J_t(\hat{P}_{t+1,t} - AV_t^t)J_{t-1}'$
-   {\tt if } $t<T$\\
-\> {\tt end}\\
-\> {\tt return }$\hat{x}_t, \hat{P}_t, \hat{P}_{t,t-1}$ {\tt for all } $t$
-\end{tabbing}
-}
-\caption{LDS Inference Algorithm}
-\vspace{0.6cm}
-\end{figure}
+\begin{algorithm}[h]
+\begin{pseudocode}
+\codename $\code{lds\_inference}\left(Y,A,C,Q,R,x,P\right)$\\
+\codeline \code{for } $t=1$ to $T$ ;forward recursion \\
+\codeline \> $x_t^{t-1} = Ax_{t-1}^{t-1}$\\
+\codeline \> $P_t^{t-1} = AP_{t-1}^{t-1}A' + Q \code{ if } t>1$\\
+\codeline \> $K_t = P_t^{t-1}C'(CP_t^{t-1}C'+R)^{-1}$\\
+\codeline \> $x_t^t = x_t^{t-1} + K_t(y_t - Cx_t^{t-1})$\\
+\codeline \> $P_t^t = P_t^{t-1} - K_tCP_t^{t-1}$ \\
+\codeline $\hat{P}_{T,T-1}=(I-K_TC)AP_{T-1}^{T-1}$ ;initialize \\
+\codeline $\code{for } t=T$ to $2$ ; backward recursion \\
+\codeline \> $J_{t-1} = P_{t-1}^{t-1}A'(P_t^{t-1})^{-1}$\\
+\codeline \> $\hat{x}_{t-1} = x_{t-1}^{t-1} + J_{t-1}(\hat{x}_{t} - Ax_{t-1}^{t-1})$\\
+\codeline \> $\hat{P}_{t-1} = P_{t-1}^{t-1} + J_{t-1}(\hat{P}_t - P_t^{t-1})J_{t-1}'$\\
+\codeline \> $\hat{P}_{t,t-1} = P_t^t J_{t-1}' + J_t(\hat{P}_{t+1,t} - AV_t^t)J_{t-1}' \code{ if } t<T$\\
+\codeline $\code{ return } \hat{x}_t, \hat{P}_t, \hat{P}_{t,t-1} \forall t$
+\end{pseudocode}
+\end{algorithm}
 
-\begin{figure}[!hbp]
-\center{
-\begin{tabbing}
-{\tt LDS}\={\tt Learn}\=$(Y,k,$\=$\epsilon)$\\ \> {\tt initialize } $A, C, Q, R,
-x_1^0, P_1^0$\\ \> $\alpha = \sum_t y_ty_t'$\\ \> {\tt while change in
-  likelihood >} $\epsilon$\\ \> \> {\tt LDSInference}$(Y,A,C,Q,R,x_1^0,P_1^0)$
-\% E-step\\ \> \> {\tt initialize } $\delta = 0, \gamma = 0, \sigma = 0$\\ \> \>
-   {\tt for } $t = 1$ {\tt to } $T$\\ \> \> \> $\delta = \delta +
-   y_t\hat{x}_t$\\ \> \> \> $\gamma = \gamma + \hat{x}_t\hat{x}_t' +
-   \hat{P}_t$\\ \> \> \> $\beta = \beta + \hat{x}_t\hat{x}_{t-1}' + P_{t,t-1}$
-       {\tt if } $t>1$\\ \> \> {\tt end}\\ \> \> $\gamma_1 = \gamma -
-       \hat{x}_T\hat{x}_T' - \hat{P}_T$\\ \> \> $\gamma_2 = \gamma -
-       \hat{x}_1\hat{x}_1' - \hat{P}_1$\\ \> \> \% M-step\\ \> \> $C = \delta
-       \gamma^{-1}$\\ \> \> $R = (\gamma - C\delta')/T$\\ \> \> $A =
-       \beta\gamma_1^{-1}$\\ \> \> $Q = (\gamma_2 - A\beta') / (T-1)$\\ \> \>
-       $x_1^0 = \hat{x}_1$\\ \> \> $P_1^0 = \hat{P}_1$\\ \> {\tt end}\\ \> {\tt
-         return } $A, C, Q, R, x_1^0, P_1^0$
-\end{tabbing}
-}
-\caption{LDS Learning Algorithm}
-\vspace{0.6cm}
-\end{figure}
+\begin{algorithm}[h]
+\begin{pseudocode}
+\codename $\code{lds\_learn}\left(Y,k,\epsilon\right)$\\
+\codeline $A, C, Q, R,x_1^0, P_1^0$\\
+\codeline $\alpha = \sum_t y_ty_t'$\\ 
+\codeline $\code{while}$ change in likelihood $> \epsilon$\\ 
+\codeline \> ; E-step \\
+\codeline \> $\code{lds\_inference}(Y,A,C,Q,R,x_1^0,P_1^0)$ \\
+\codeline \> $\delta = 0, \gamma = 0, \sigma = 0$ \\
+\codeline \> $\code{for} t = 1$ to $T$\\ 
+\codeline \> \> $\delta = \delta + y_t\hat{x}_t$\\ 
+\codeline \> \> $\gamma = \gamma + \hat{x}_t\hat{x}_t' + \hat{P}_t$\\ 
+\codeline \> \> $\beta = \beta + \hat{x}_t\hat{x}_{t-1}' + P_{t,t-1} \code{ if } t>1$\\ 
+\codeline \> $\gamma_1 = \gamma - \hat{x}_T\hat{x}_T' -\hat{P}_T$\\ 
+\codeline \> $\gamma_2 = \gamma - \hat{x}_1\hat{x}_1' - \hat{P}_1$\\ 
+\codeline \> ; M-step\\ 
+\codeline \> $C = \delta \gamma^{-1}$\\ 
+\codeline \> $R = (\gamma - C\delta')/T$\\ 
+\codeline \> $A = \beta\gamma_1^{-1}$\\ 
+\codeline \> $Q = (\gamma_2 - A\beta') / (T-1)$\\ 
+\codeline \> $x_1^0 = \hat{x}_1$\\ 
+\codeline \> $P_1^0 = \hat{P}_1$\\ 
+\codeline $\code{return } A, C, Q, R, x_1^0, P_1^0$
+\end{pseudocode}
+\end{algorithm}
 
@@
 
-\begin{figure}[!hbp]
-\center{
-\begin{tabbing}
-{\tt ob}\={\tt ser}\={\tt vation}\={\tt Genera}\={\tt tion}$(A,C,R,Q,\mu,\Sigma,T,N)$\\
-\> $mu_x = mu[T-1];$\\
-\> $sigma_x = sigma[T-1];$\\
-\> {\tt for} $i = 1$ {\tt to } $N$\\
-\> \> $mu_y = C A mu_x;$\\
-\> \> $sigma_y = CA sigma_x A'C' + CQC' + R;$\\
-\> \> $o[i] = $ {\tt gaussianNumberGeneration}$(mu_y, sigma_y);$\\
-\> \> $mu_x = A mu_x$\\
-\> \> $sigma_x = A  sigma_x A + Q$\\
-\> {\tt end}\\
-\> {\tt return} $o;$
-\end{tabbing}
-}
-\caption{Generating Observations from an LDS Algorithm}
-\vspace{0.6cm}
-\end{figure}
+\begin{algorithm}[h]
+\begin{pseudocode}
+\codename $\code{observation\_generation}\left(A,C,R,Q,\mu,\Sigma,T,N\right)$\\
+\codeline $\mu_x = \mu[T-1];$\\
+\codeline $\sigma_x = \sigma[T-1];$\\
+\codeline $\code{for } i = 1$ to $N$\\
+\codeline \> $\mu_y = C A \mu_x;$\\
+\codeline \> $\sigma_y = CA \sigma_x A'C' + CQC' + R;$\\
+\codeline \> $o[i] = \code{gaussian\_number\_generation}(\mu_y, \sigma_y)$\\
+\codeline \> $\mu_x = A \mu_x$\\
+\codeline \> $\sigma_x = A  \sigma_x A + Q$\\
+\codeline $\code{return } o$
+\end{pseudocode}
+\end{algorithm}

### 2015-07-27 books/neural-networks-and-deep-learning/src/expand_mnist.py

@@ -26,10 +26,10 @@ import numpy as np
 
 print("Expanding the MNIST training set")
 
-if os.path.exists("../data/mnist_expanded.pkl.gz"):
+if os.path.exists("/home/burak/Downloads/mnist_expanded.pkl.gz"):
     print("The expanded training set already exists.  Exiting.")
 else:
-    f = gzip.open("../data/mnist.pkl.gz", 'rb')
+    f = gzip.open("/home/burak/Downloads/mnist.pkl.gz", 'rb')
     training_data, validation_data, test_data = cPickle.load(f)
     f.close()
     expanded_training_pairs = []
@@ -55,6 +55,6 @@ else:
     random.shuffle(expanded_training_pairs)
     expanded_training_data = [list(d) for d in zip(*expanded_training_pairs)]
     print("Saving expanded data. This may take a few minutes.")
-    f = gzip.open("../data/mnist_expanded.pkl.gz", "w")
+    f = gzip.open("/home/burak/Downloads/mnist_expanded.pkl.gz", "w")
     cPickle.dump((expanded_training_data, validation_data, test_data), f)
     f.close()

### 2015-07-28 ms-thesis/ver2/kalman/ms-thesis-kalman.tex
-

@@
 as follows:
-\begin{eqnarray}
+\begin{eqnarray*}
 x_{t+1} = Ax_t + v_t \mlabel{1}\\
 y_t = Cx_t + w_t \mlabel{2}
-\end{eqnarray}
+\end{eqnarray*}
 where $w_0,w_1,...$ are white Gaussian noise with $\mathbf{E}w_t = 0$ and $cov
@@
 We can also re-write this for conditional and with the mean;
-\begin{eqnarray}
+\begin{eqnarray*}
 \left[ \begin{array}{c}
@@
       \end{array} \right] \mlabel{3}
-\end{eqnarray}
+\end{eqnarray*}
 Same for the joint distribution for $x_{t+1}, y_{t+1}$.
-\begin{eqnarray}
+\begin{eqnarray*}
 \left[ \begin{array}{c}
@@
       \end{array} \right] \mlabel{4}
-\end{eqnarray}
+\end{eqnarray*}
 
@@
 
-\begin{eqnarray}
+\begin{eqnarray*}
 \mu = 
@@
       \end{array} \right] \mlabel{6}
-\end{eqnarray}
+\end{eqnarray*}
 \begin{eqnarray*}
@@
 get {\em conditioned} $\mu$ and $\Sigma$, as
-\begin{eqnarray}
+\begin{eqnarray*}
 \mu_{1|2} &=& \mu_1 + \Sigma_{12}\Sigma_{22}^{-1}(x_2 - \mu_2) \mlabel{5} \\
 \Sigma_{1|2} &=& \Sigma_{11}-\Sigma_{12}\Sigma_{22}^{-1}\Sigma_{21} \nonumber
-\end{eqnarray}
+\end{eqnarray*}
 Now plugging elements from  (5) into (4) according
@@
 If we declare $K_t \equiv P_{t+1}^tC' (CP_{t+1}^tC'+ \Sigma_{w})^{-1}$ then,
-\begin{eqnarray}
+\begin{eqnarray*}
 \hat{x}_{t+1}^{t+1} &=& \hat{x}_{t+1}^t + K_t (y_{t+1} -
@@
 P_{t+1}^{t+1} &=& P_{t+1}^t - K_t C P_{t+1}^t
-\end{eqnarray}
+\end{eqnarray*}
 
@@
 we can drop $x_{t+1}$ from LHS.
-\begin{eqnarray}
+\begin{eqnarray*}
 \hat{x}_{t}' & \equiv & \mathbf{E}[x_t|y_0,...,y_T] \nonumber\\
@@
 & = & \hat{x}_{t}^t +L_t(\hat{x}_{t+1}' -\hat{x}_{t+1}^t) \label{xsmooth}
-\end{eqnarray}
+\end{eqnarray*}
 Using similar algebra, we get
@@
 Given,
-\begin{eqnarray}
+\begin{eqnarray*}
 x_{t+1} = Ax_t + v_t \\
 y_t = Cx_t + w_t
-\end{eqnarray}
+\end{eqnarray*}
 where $v_0,v_1,...$ is white Gaussian noise with $\mathbf{E}v_t = 0$ and $cov \:
@@
 0$ and $cov \: w_t = R$, we can determine $y_{t+1}$ as
-\begin{eqnarray}
+\begin{eqnarray*}
 y_{t+1} &=& Cx_{t+1} + w_{t+1}\\
@@
 &=& CAx_t + Cv_t + w_{t+1}
-\end{eqnarray}
+\end{eqnarray*}
 which gives us the distribution of the emission Gaussian $y_{t+1}$. The
 conditional expectation and variance of this distribution given $y_{1,..,t}$ is
-\begin{eqnarray}
+\begin{eqnarray*}
 y_{t+1} &=& Cx_{t+1} + w_{t+1}\\
@@
 cov(y_{t+1}|y_{0,...,t}) &=& CA P_{t|t}A'C' + CQC' + R
-\end{eqnarray}
+\end{eqnarray*}
 We can apply this recursive formula to have our KF generate $M$ many random

### 2015-07-28 ms-thesis/ver2/cover/ms-thesis-cover.tex
-

@@
-\title{FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN
-  MARKOV MODELS \\ \\ HMM VE KALMAN FILTRELERI KULLANARAK FINANSAL ZAMAN SERISI TAHMINI}
+\chapter{Introduction}
 
-\degree{B.E., Computer Engineering, Stevens Institute of Technology, 1996}
+FINANCIAL TIME SERIES PREDICTION USING KALMAN FILTERS AND HIDDEN MARKOV MODELS
 
-\author{Burak Bayramli}
+Burak Bayramli
 
-\program{Systems and Control Engineering}
+B.E., Computer Engineering, Stevens Institute of Technology, 1996
 
-\supervisor{Prof. Fikret Gurgen}
+Systems and Control Engineering
 
-\examineri{Prof. Levent Akin}
+Prof. Fikret Gurgen
 
-\examinerii{Prof. Ethem Alpaydin}
+Prof. Levent Akin
 
-\dateofapproval{..............................}
+Prof. Ethem Alpaydin
 
-\begin{document}
-
-\pagenumbering{roman}
-\makemstitle        % For M.S. theses
-
-\makeapprovalpage
-
-\begin{acknowledgements}
 I would like to thank my thesis advisor Dr. Gürgen for his help, direction and
@@
 
-Thanks to Dr. Ethem Alpaydın for first suggesting Systems and Control
+Thanks to Dr. Ethem Alpaydin for first suggesting Systems and Control
 Engineering department as a good master's goal, and second for all the knowledge
@@
 
-I extend warmest thanks to Dr. Levent Akın for his excellent {\em Autonomous
+I extend warmest thanks to Dr. Levent Akin for his excellent {\em Autonomous
 Robotics} class and his suggestion that I look into {\em Probabilistic
@@
 
-I would also like to thank my parents Ercan and Füsun Bayramlı for their support
+I would also like to thank my parents Ercan and Fusun Bayramli for their support
 without whom none of this would have been possible.
-\end{acknowledgements}
 
-\begin{abstract}
+\newpage
+
 Financial markets are challenging targets for analysis and prediction. The
@@
 margin. We were glad to see that this was indeed the case.
-\end{abstract}
 
-\begin{ozet}
-Finans piyasaları, analiz ve fiyat tahmini yapmak açısından zor bir
-alandır. Tahmin yapılmasını zorlaştıran faktör, herhangi bir gündeki fiyatın
-ortaya çıkmasında çok fazla etkinin olmasıdır. Tahmin ve analiz için şimdiye
-kadar pek çok istatistiki ve diğer yöntemler bu amaç için seferber edilmiştir -
-biz de bu yolda bir adım atmaya uğraştık.
-
-Bu bağlamda Gizli Markov Modelleri ve Kalman Filtreleri adlı iki kuvvetli
-metotun, geçmiş senet verisini modelleyebilerek geleceğe dönük tahmin
-yapabileceğini varsay\-dık, ve iki metodu birleştirerek KMM adlı üçüncü bir
-metodun performansına baktık. Fiyat tahmin sayılarıni üretirken Monte Carlo
-metotu kullanarak uç noktalarda olan tahminleri ortalama içinde düzeltmeye ve
-böylece daha iyi bir tahmine ulaşmaya çabaladık. Model eğitildikten sonra onu
-ileriye dönük tahminlerde kullanmak için ise, GMM durumunda Viterbi algoritması
-ile geçiş olasılıklarını rasgele sayı üreterek takip eden bir algoritma, Kalman
-Filtreleri durumunda ise en son gelinen zaman noktasından bir sonrakine geçiş
-için bizim değiştirdiğimiz zaman güncelleme denklemleri, ve yine rasgele sayı
-üreten bir algoritma kullandık.
-
-Bu tezin bir diğer amacı tarif ettiğimiz metotların diğer klasik metotlara
-kıyasla nasıl başarı göstereceğini ortaya koymaktı. Bunun için Yapay Sinir
-Ağları, polinom regresyon ve en son pür rasgele sayı üreterek tahmin yapmaya
-uğraşan üç diğer metotu test senaryolarımıza dahil ettik. Bu metotların arasında
-rasgele üretim en alt seviyede performans göstermesini beklenilen metot idi,
-diğer tüm metotlar farklı derecelerde bu metotu geçmeliydi, ki deneyler
-sonucunda durumun böyle olduğunu gözlemledik.
-\end{ozet}
+\newpage
 
@@
 \listoftables
-
-\begin{symabbreviations}
-% The title will be typeset as "LIST OF SYMBOLS/ABBREVIATIONS".
-% If you prefer it as "LIST OF SYMBOLS" or "LIST OF ABBREVIATIONS"
-% use the environments "symbols" or "abbreviations", respectively.
-%
-% Use a separate \sym command for each symbols definition.
-% First Latin symbols in alphabetical order
-
-\sym{$a_{ij}$}{HMM transition probability $p(q_t^j=1 | q_t^i=1)$ of hidden states}
-\sym{$A$}{HMM hidden transitions matrix. $a_{ij}$ accesses elements of $A$}
-\sym{$A$}{Linear multiplier for KF time transition from time $t$ to $t+1$}
-\sym{$C$}{A vector holding all means of emission Gaussians for HMM}
-\sym{$C$}{Linear multiplier for KF that translates $x_t$  to $y_t$}
-\sym{$h_t$}{Sum of errors going back $p$ days for ARCH / GARCH / EGARCH modeling.}
-\sym{$K_t$}{Kalman gain matrix at time $t$} 
-\sym{$P_t$}{Price of a security at time $t$}
-\sym{$R_t$}{Rate of return for financial time series data}
-\sym{$R$}{Variance of the emission Gaussian for HMM}
-\sym{$P_{t}^{t}$}{Conditional variance of $x$  at time $t$ given output
-emissions $y_0,...,y_t$}
-\sym{$P_{t,t-1}$}{Joint covariance necessary for KF lag-one covariance smoothing
-calculations} 
-\sym{$q_t$}{HMM's hidden state symbol at time $t$}
-\sym{$T$}{Number of training data points} 
-\sym{$v_t$}{White noise - Gaussian distribution with zero mean and covariance $Q$}
-\sym{$\hat{x}_{t}^t$}{Conditional expectation of $x$ at time $t$ given output
-emissions $y_0,...,y_t$} 
-\sym{$x_t$}{Random variable representing hidden state at time $t$ for KF
-(Gaussian distribution).} 
-\sym{$y$}{All of the measured/observed data for HMM, KF and KMM}
-\sym{$w_t$}{White noise - Gaussian distribution with zero mean and covariance $R$}
-
-\\\\
-
-\sym{$\alpha(q_t)$}{Probability of emitting partial sequence of outputs
-$y_0,...,y_t$ and ending up in state $q_t$ \cite{jordan} for HMM}
-\sym{$\beta(q_t)$}{Probability of emitting partial sequence of outputs
-$y_{t+1},..,y_T$ given that system starts in state $q_t$ for HMM}
-\sym{$\eta$}{Distribution representing output emissions for HMM}
-\sym{$\epsilon$}{Error}
-\sym{$\gamma_{t}$}{HMM posterior probability $p(q_t|y)$ at time $t$}
-\sym{$\pi$}{HMM initial probability distribution}
-\sym{$\xi_t$}{HMM cooccurence probabilities of $q_t$ and $q_{t+1}$}
-
-\\\\
-
-\sym{ANN}{Artificial Neural Network}
-\sym{ARCH}{Autoregressive conditional heteroskedasticity}
-\sym{DJI}{Dow Jones Industrial}
-\sym{EM}{Expectation Maximization}
-\sym{GARCH}{General ARCH}
-\sym{HMM}{Hidden Markov Models}
-\sym{HME}{Hidden Markov Experts}
-\sym{iid}{Independent Identically Distributed}
-\sym{ISE}{Istanbul Stock Exchange}
-\sym{KF}{Kalman Filters} 
-\sym{KMM}{A hybrid method combining KF and HMM} 
-\sym{MC}{Markov Chains} 
-\sym{MLP}{Multilayer perceptron} 
-\sym{RTS}{Rauch-Tung-Striebel} 
-\sym{SSM}{State space model}
-
-\end{symabbreviations}

### 2015-07-28 ms-thesis/ver2/build.py

@@ -2,6 +2,7 @@ import os, sys
 
 if len(sys.argv) == 1 or sys.argv[1] == 'tex':
     os.system("pdflatex -shell-escape ms-thesis.tex")
+    os.system("cp ms-thesis.pdf $HOME/Dropbox/Public/skfiles/")
     os.system("evince ms-the*.pdf")
     exit()
     

### 2015-07-29 books/Quantitative_Trading_Chan/calculateMaxDD.py
-

@@ -1,18 +1,0 @@
-from scipy import io as spio
-import numpy as np
-def calculateMaxDD(cumret):
-    highwatermark=np.zeros(len(cumret))
-    drawdown=np.zeros(len(cumret))
-    drawdownduration=np.zeros(len(cumret))
-    for t in range(1,len(cumret)):
-        highwatermark[t]=np.max([highwatermark[t-1], cumret[t]])
-        drawdown[t]=(1+cumret[t])/(1+highwatermark[t])-1
-        if (drawdown[t]==0):
-            drawdownduration[t]=0
-        else:
-            drawdownduration[t]=drawdownduration[t-1]+1
-    return np.min(drawdown), np.max(drawdownduration)
-
-a = spio.loadmat('spy_tmp')['a'][0]
-print calculateMaxDD(a)
-

### 2015-07-29 books/Quantitative_Trading_Chan/read_mat.py
-

@@ -1,0 +1,4 @@
+# to save a matlab matrix
+# save('dosya','A')
+from scipy import io as spio
+a = spio.loadmat('dosya')['A'][0]

### 2015-08-23 books/Advanced_Time_Series_and_Forecasting_Hansen/gdp2.r
-

@@ -1,0 +1,171 @@
+# gdp2.r #
+
+library(quadprog)		# You may need to install the package #
+library(tseries)		# You may need to install the package #
+
+gdpdata <- read.table("gdpdata.txt")
+gdpdata <- as.matrix(gdpdata)
+n=nrow(gdpdata)
+year=as.matrix(gdpdata[1:n,1])
+quarter=as.matrix(gdpdata[1:n,2])
+t=year+(quarter-1)/4
+
+gdp=as.matrix(gdpdata[1:n,3])
+gdpg=as.matrix(gdpdata[1:n,4])
+t10=as.matrix(gdpdata[1:n,5])
+t3=as.matrix(gdpdata[1:n,6])
+aaa=as.matrix(gdpdata[1:n,7])
+baa=as.matrix(gdpdata[1:n,8])
+hs=as.matrix(gdpdata[1:n,9])
+bp=as.matrix(gdpdata[1:n,10])
+spread=t10-t3
+dspread=baa-aaa
+
+# Create Data Matrix Using 12 Initial Conditions @
+kk=12			# Number of initial conditions #
+nn=n-kk			# Number of data points less number of initial conditions equals number of observations #
+y=as.matrix(gdpg[(1+kk):n])	# dependent variable #
+x=matrix(1,nn+1,1)	# Regressors, first column (ones), one more observation than dependent variable (for forecast) #
+for (j in 1:kk)
+{ x=cbind(x,gdpg[(1+kk-j):(n-j+1)]) } 	# X matrix columns, lags of y #
+
+
+# Model Combination #
+kn=kk+1				# Number of models = AR(0) through AR(kk) #
+yf=matrix(0,kn,1)		# vector of forecasts (empty for now) #
+ee=matrix(0,nn,kn)		# matrix of prediction errors (empty for now) #
+
+for (k in 1:kn)
+{
+  xk=x[1:nn,1:k]
+  xf=x[nn+1,1:k]
+  xxi=solve(t(xk)%*%xk)
+  beta=xxi%*%(t(xk)%*%y)
+  e=y-xk%*%beta
+  h=rowSums((xk%*%xxi)*xk)
+  eh=e/(1-h)
[…]

### 2015-08-23 books/Advanced_Time_Series_and_Forecasting_Hansen/rates1.r

@@ -1,0 +1,98 @@
+# rates1.r #
+
+rates <- read.table("rates.txt")
+rates <- as.matrix(rates)
+n=nrow(rates)
+year=as.matrix(rates[1:n,1])
+month=as.matrix(rates[1:n,2])
+t=year+(month-1)/12
+r=as.matrix(rates[1:n,4])
+
+plot(t,r,main="10-Year Treasury Bill Rate",type="l",xlab="",ylab="")
+
+# First Difference #
+dr=r[2:n]-r[1:(n-1)]
+windows()
+plot(t[2:n],dr,main="Change in 10-Year Rate",type="l",xlab="",ylab="")
+n=n-1
+
+# Create Data Matrix Using 24 Initial Conditions @
+kk=24			# Number of initial conditions #
+nn=n-kk			# Number of data points less number of initial conditions equals number of observations #
+y=as.matrix(dr[(1+kk):n])	# dependent variable #
+x=matrix(1,nn+1,1)	# Regressors, first column (ones), one more observation than dependent variable (for forecast) #
+for (j in 1:kk)
+{ x=cbind(x,dr[(1+kk-j):(n-j+1)]) } 	# X matrix columns, lags of y #
+
+
+# Estimate AR(12) #
+k=13			# Number of lags + 1 for intercept #
+xk=x[1:nn,1:k]		# X matrix, for estimation	#
+xf=x[nn+1,1:k]		# X vector, for forecast	#
+xxi=solve(t(xk)%*%xk)	# X'X #
+beta=xxi%*%(t(xk)%*%y)	# OLS #
+e=y-xk%*%beta		# OLS residual #
+h=rowSums((xk%*%xxi)*xk)	# leverage values #
+eh=e/(1-h)			# LOO prediction residuals #
+xe=xk*(eh%*%matrix(1,1,k))	# x*e #
+omega=t(xe)%*%xe		# Central part of Variance matrix #
+v=xxi%*%omega%*%xxi		# variance matrix estimator #
+se=sqrt(diag(v))		# Standard errors #
+betas=cbind(beta,se)		# estimates + standard errors #
+yf=xf%*%beta			# Point forecast = Estimated change #
+print("Last 4 values, level and change")
[…]

### 2015-08-23 books/Advanced_Time_Series_and_Forecasting_Hansen/rates2.r

@@ -1,0 +1,81 @@
+# rates2.r #
+
+library(quadprog)		# You may need to install the package #
+library(tseries)		# You may need to install the package #
+
+rates <- read.table("rates.txt")
+rates <- as.matrix(rates)
+n=nrow(rates)
+year=as.matrix(rates[1:n,1])
+month=as.matrix(rates[1:n,2])
+t=year+(month-1)/12
+r=as.matrix(rates[1:n,4])
+dr=r[2:n]-r[1:(n-1)]
+n=n-1
+
+# Create Data Matrix Using 24 Initial Conditions @
+kk=24			# Number of initial conditions #
+nn=n-kk			# Number of data points less number of initial conditions equals number of observations #
+y=as.matrix(dr[(1+kk):n])	# dependent variable #
+x=matrix(1,nn+1,1)	# Regressors, first column (ones), one more observation than dependent variable (for forecast) #
+for (j in 1:kk)
+{ x=cbind(x,dr[(1+kk-j):(n-j+1)]) } 	# X matrix columns, lags of y #
+
+
+# Model Combination #
+kn=kk+1				# Number of models = AR(0) through AR(kk) #
+yf=matrix(0,kn,1)		# vector of forecasts (empty for now) #
+ee=matrix(0,nn,kn)		# matrix of prediction errors (empty for now) #
+
+for (k in 1:kn)
+{
+  xk=x[1:nn,1:k]
+  xf=x[nn+1,1:k]
+  xxi=solve(t(xk)%*%xk)
+  beta=xxi%*%(t(xk)%*%y)
+  e=y-xk%*%beta
+  h=rowSums((xk%*%xxi)*xk)
+  eh=e/(1-h)
+  yf[k]=xf%*%beta  
+  ee[,k]=eh	
+}
+
+Dmat=(t(ee)%*%ee)/nn
+dvec=matrix(0,kn,1)
+Amat=t(rbind(matrix(1,1,kn),diag(kn)))
+bvec=rbind(1,matrix(0,kn,1))
+QP <- solve.QP(Dmat,dvec,Amat,bvec,bvec)
+w <- QP$solution
+w <- as.matrix(w)
+e=ee%*%w
+cv=t(w)%*%Dmat%*%w
+
+yff=t(yf)%*%w
+print("Models, Weights")
+print(cbind(seq(0,kk),w))
+print("CV, Combination Forecast, Level Forecast")
+print(cbind(cv,yff,r[n+1]+yff))
+
[…]

### 2015-08-23 books/Advanced_Time_Series_and_Forecasting_Hansen/gdp2.r
-

@@ -165,7 +165,7 @@ print(like)
 t1 <- as.matrix(t[(kk+1):n])
 e2=e^2
 plot(t1,e,main="Leave-One-Out Prediction Residuals",type="l",xlab="",ylab="")
-windows()
+X11()
 plot(t1,e2,main="Squared Prediction Residuals",type="l",xlab="",ylab="")
-windows()
+X11() 
 plot(t1,var,main="Estimated Variance",type="l",xlab="",ylab="")

### 2015-08-23 books/Advanced_Time_Series_and_Forecasting_Hansen/rates1.r

@@ -12,7 +12,7 @@ plot(t,r,main="10-Year Treasury Bill Rate",type="l",xlab="",ylab="")
 
 # First Difference #
 dr=r[2:n]-r[1:(n-1)]
-windows()
+X11()
 plot(t[2:n],dr,main="Change in 10-Year Rate",type="l",xlab="",ylab="")
 n=n-1
 

### 2015-08-23 books/Advanced_Time_Series_and_Forecasting_Hansen/rates2.r

@@ -75,7 +75,7 @@ print(like)
 t1 <- as.matrix(t[(kk+1):n])
 e2=e^2
 plot(t1,e,main="Leave-One-Out Prediction Residuals",type="l",xlab="",ylab="")
-windows()
+X11()
 plot(t1,e2,main="Squared Prediction Residuals",type="l",xlab="",ylab="")
-windows()
+X11()
 plot(t1,var,main="Estimated Variance",type="l",xlab="",ylab="")

### 2015-08-24 quand.py
-

@@ -1,0 +1,2 @@
+# BwTLp2sKeKsqtLJZpCQc
+

### 2015-08-24 yahoo.py

@@ -1,0 +1,14 @@
+import datetime
+import pandas as pd
+import pandas.io.data
+
+# Yahoo Finance
+#aapl = pd.io.data.get_data_yahoo('MSFT', 
+#                                 start=datetime.datetime(2006, 10, 1), 
+#                                 end=datetime.datetime(2015, 6, 1))
+#aapl.to_csv('/tmp/aapl_ohlc.csv')
+
+# FRED
+jpy = pd.io.data.DataReader('DEXUSEU', 'fred')
+jpy.to_csv('/tmp/DEXJPUS.csv')
+

### 2015-08-25 books/Logic_of_Political_Survival_Mesquita/code/ch41.tex
-

@@
 
+%%%%%%%% Run these first %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{minted}[fontsize=\footnotesize]{python}
+import pandas as pd
+df = pd.read_csv('../bdm2s2_nation_year_data_may2002.csv')
+print list(df.columns)
+\end{minted}
+
+\begin{verbatim}
+['Unnamed: 0', 'year', 'AGEA', 'AGEI', 'AUT', 'BRITCOL', 'CATH', 'COMEX',
+'EDT', 'EDTG', 'ELF60', 'ETHNIC', 'GINI', 'GXPDEDUC', 'GXPDHLTH',
+'GXPDSSEC', 'GXPKTOTL', 'INEQ', 'INST', 'LEGSELEC', 'MOSLEM', 'OIL',
+'PROT', 'REGION', 'REGION2', 'RELIGION', 'RIOTS', 'STRIKES', 'UNSTABLE',
+'WAR', 'ccode', 'TAXGDP', 'TAXGNP', 'TREV', 'PRIVCON', 'INCOMEY',
+'INCOMET', 'DIRT', 'INDIRT', 'democ', 'autoc', 'polity', 'polity2',
+'durable', 'xrreg', 'xrcomp', 'xropen', 'xconst', 'parreg', 'parcomp',
+'exrec', 'exconst', 'polcomp', 'bdate', 'bprec', 'change', 'd3', 'sf',
+'regtrans', 'country', 'BANKScode', 'Radios', 'Newspapers', 'Assinations',
+'GenStrikes', 'GovCrises', 'Purges', 'Riots', 'Revolutions', 'AntiGovDem',
+'ConflictIndex', 'RegimeType', 'Coups', 'HeadofState', 'Guerilla',
+'RegisteredVoters', 'MajConstChg', 'Premier', 'Legselec', 'NatGovRevenue',
+'NatGovExpend', 'defenseovertotal', 'name', 'chcode', 'longname',
+'coding_differs', 'WorldBankCode', 'WorldBankName', 'fatlev1', 'hostlev1',
+'legselec', 'esolow', 'bhkavg', 'exchprem', 'sh_staa', 'sh_med_b',
+'sh_med_p', 'sh_h2o_s', 'immig', 'emig', 'ID', 'NAME', 'PR', 'CL', 'cap',
+'milper', 'milex', 'energy', 'irst', 'majpow', 'warnoEX', 'EXdeaths',
+'EXintside', 'EXinitiate', 'sysstat', 'EX_war_start', 'EXduration',
+'EX_WAR', 'number_of_EXwars', 'biggestEXwar', 'warno', 'duration',
+'deaths', 'outcome', 'initiate', 'is_war_start', 'IS_WAR',
+'number_of_ISwars', 'biggestISwar', 'warnoCL', 'CLdeaths', 'CLintside',
+'CLinitiate', 'CL_war_start', 'CLduration', 'CL_WAR', 'number_of_CLwars',
+'biggestCLwar', 'ai', 'sd', 'ainew', 'sdnew', 'idgurr', 'worldbankcode',
+'WB_pop', 'WB_gdppc_constUS95', 'WB_growth', 'WB_gdp_curUS', 'WB_gdp95',
+'Aid_pcUS', 'FemSec', 'Doctors', 'MilPerson', 'MilEx', 'MilExGNI',
+'InfMort', 'LowBirth', 'LifeExp', 'Labor_PrimEd', 'ImmuneMEASLES',
+'ImmuneDPT', 'Illiteracy', 'Beds', 'HealthEXP', 'HealthExpPPP',
+'GrossNatSaving', 'GrossDomSaving', 'Capital_formation', 'Saving_dom',
+'FDI', 'Foreign_Finnace', 'ConsumptionSpending', 'Expenditure',
+'Deathrate', 'newspaper', 'HealthExpend', 'EducExpend',
+'CapitalExpend_tot', 'CurRevenue', 'GrossDomInvest', 'GovDebt',
+'Birthrate', 'total_aid', 'aid_cap_constUS', 'aid_gdp', 'CapitalExpend',
+'pop', 'rgdpch', 'rgdpl', 'c', 'i', 'g', 'rgdptt', 'y', 'cgdp', 'cc', 'ci',
+'cg', 'p', 'pc', 'pi', 'pg', 'xr', 'rgdpea', 'rgdpw', 'kapw', 'kdur',
+'knres', 'kother', 'kres', 'ktranp', 'open', 'rgnp', 'ipri', 'stliv',
+'finittrm', 'yrcurnt', 'military', 'auton', 'eda', 'strikes', 'nspolpr',
+'agovdem', 'govtcris', 'corrupti80_89', 'purges', 'coups', 'riots',
+'constchg', 'cabchg', 'genocide', 'worldbankname', 'NatYr', 'cowreg',
+'regyr', 'demaut', 'lpop', 'lpop_WB', 's', 'S', 'W', 'W_extend', 'W_mod',
+'polityW', 'polityWn', 'Parl_Pres', 'w2', 'demaut2', 'WoverS', 'dbigw',
+'dW', 'dw30', 'lagW', 'lagW2', 'dwos30', 'grow', 'LongGR1', 'LongGR2',
+'LongGR3', 'LongGR4', 'lagGrow', 'lrgdpc', 'laglrgdpc', 'lgdpWB',
+'laglgdpWB', 'VarGrow', 'VarGrow_', 'build', 'bal_bud', 'VarBud',
+'Var_bud', 'Klepto', 'nexcap'] 
+\end{verbatim}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+%load_ext rpy2.ipython
+%R library(lme4)
+\end{minted}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+df['polchange'] = (df.W-df.shift(2).W)**2
+df['DW20'] = df.W-df.shift(2).W
+df['Klepto'] = (df['TAXGDP']-df['Expenditure']).abs()
+\end{minted}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+import statsmodels.formula.api as smf
+results = smf.ols('WB_growth ~ W + S + DW20 + polchange', data=df).fit()
+print results.summary()
+\end{minted}
+
+\begin{verbatim}
+                            OLS Regression Results                            
+==============================================================================
+Dep. Variable:              WB_growth   R-squared:                       0.003
+Model:                            OLS   Adj. R-squared:                  0.003
+Method:                 Least Squares   F-statistic:                     4.266
+Date:                Tue, 25 Aug 2015   Prob (F-statistic):            0.00190
+Time:                        16:41:42   Log-Likelihood:                -16242.
+No. Observations:                4985   AIC:                         3.249e+04
+Df Residuals:                    4980   BIC:                         3.253e+04
+Df Model:                           4                                         
+Covariance Type:            nonrobust                                         
+==============================================================================
+                 coef    std err          t      P>|t|      [95.0% Conf. Int.]
+------------------------------------------------------------------------------
+Intercept      4.2384      0.259     16.382      0.000         3.731     4.746
+W             -0.3539      0.364     -0.972      0.331        -1.068     0.360
+S             -0.3540      0.331     -1.069      0.285        -1.003     0.295
+DW20           0.6777      0.620      1.093      0.274        -0.538     1.893
+polchange     -3.8690      1.014     -3.817      0.000        -5.856    -1.882
+==============================================================================
+Omnibus:                     1170.985   Durbin-Watson:                   1.341
+Prob(Omnibus):                  0.000   Jarque-Bera (JB):            61520.180
+Skew:                           0.155   Prob(JB):                         0.00
+Kurtosis:                      20.207   Cond. No.                         16.9
+==============================================================================
+
+\end{verbatim}
+
 \begin{minted}[fontsize=\footnotesize]{python}
@@
 df = pd.read_csv('bdm2s2_nation_year_data_may2002.csv')
-df['Klepto'] = (df['TAXGDP']-df['Expenditure']).abs()
 df = df[['regyr','ccode','aid_gdp','lrgdpc','pop','year','W','S','Klepto','laglrgdpc']]
@@
 \end{verbatim}
-
-\begin{minted}[fontsize=\footnotesize]{python}
-%load_ext rpy2.ipython
-%R library(lme4)
-\end{minted}
 

### 2015-08-28 books/Logic_of_Political_Survival_Mesquita/code/ch41.tex
-

--- 2015-04-09 books/Logic_of_Political_Survival_Mesquita/code/ch41.tex
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+Growth and lagged change in W
+
 \begin{minted}[fontsize=\footnotesize]{python}
 import pandas as pd
-df = pd.read_csv('bdm2s2_nation_year_data_may2002.csv')
+df = pd.read_csv('../bdm2s2_nation_year_data_may2002.csv')
+df = df.sort_index(by=['ccode','year'],ascending=True)
+\end{minted}
+
+If ccode at t-2 is different for current, than delta W for this instance
+makes no sense, drop them.
+
+\begin{minted}[fontsize=\footnotesize]{python}
+df2 = df.copy()
+df2['polchange'] = (df2.W-df2.shift(2).W)**2
+df2['DW20'] = df2.W-df2.shift(2).W
+df2['dummy'] = df2.ccode-df2.shift(2).ccode
+df2 = df2[df2.dummy == 0]
+print len(df2)
+\end{minted}
+
+\begin{verbatim}
+19059
+\end{verbatim}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+import statsmodels.formula.api as smf
+results = smf.ols('WB_growth ~ W + S + DW20 + polchange + np.log(pop)', data=df2).fit()
+print results.summary()
+\end{minted}
+
+\begin{verbatim}
+                            OLS Regression Results                            
+==============================================================================
+Dep. Variable:              WB_growth   R-squared:                       0.005
+Model:                            OLS   Adj. R-squared:                  0.003
+Method:                 Least Squares   F-statistic:                     3.315
+Date:                Tue, 25 Aug 2015   Prob (F-statistic):            0.00545
+Time:                        20:18:07   Log-Likelihood:                -10866.
+No. Observations:                3457   AIC:                         2.174e+04
+Df Residuals:                    3451   BIC:                         2.178e+04
+Df Model:                           5                                         
+Covariance Type:            nonrobust                                         
+===============================================================================
+                  coef    std err          t      P>|t|      [95.0% Conf. Int.]
+-------------------------------------------------------------------------------
+Intercept       4.6960      0.550      8.532      0.000         3.617     5.775
+W               0.3292      0.382      0.861      0.390        -0.421     1.079
+S              -0.7225      0.341     -2.121      0.034        -1.390    -0.055
+DW20            1.8791      0.646      2.910      0.004         0.613     3.145
+polchange      -2.4816      1.047     -2.371      0.018        -4.534    -0.429
+np.log(pop)    -0.0261      0.056     -0.468      0.640        -0.136     0.083
+==============================================================================
+Omnibus:                      408.738   Durbin-Watson:                   1.512
+Prob(Omnibus):                  0.000   Jarque-Bera (JB):             4176.816
+Skew:                          -0.006   Prob(JB):                         0.00
+Kurtosis:                       8.385   Cond. No.                         99.7
+==============================================================================
+
+Warnings:
+[1] Standard Errors assume that the covariance matrix of the errors is correctly specified.
+\end{verbatim}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{minted}[fontsize=\footnotesize]{python}
+%load_ext rpy2.ipython
+%R library(lme4)
+\end{minted}
+
+\begin{minted}[fontsize=\footnotesize]{python}
+import pandas as pd
 df['Klepto'] = (df['TAXGDP']-df['Expenditure']).abs()

### 2015-08-29 dsync.sh
-

@@ -1,10 +1,12 @@
-disk="burak/New Volume"
-#disk="burak/17EA-3758"
-echo "Copying files to /media/$disk/archive"
-rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/kitaplar "/media/$disk/archive"
-rsync  -W --size-only  --delete-after -P -avzh /home/burak/kod "/media/$disk/archive" 
-rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/classnotes "/media/$disk/archive" 
-rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/emacs-ipython "/media/$disk/archive" 
-rsync  -W --size-only  --delete-after -P -avzh /home/skorsky/mindmeld "/media/$disk/archive" 
-rsync  -W --size-only  --delete-after -P -avzh /home/burak/Dropbox "/media/$disk/archive" 
-#rsync  -W --size-only  --delete-after -P -avzh /home/burak/Music "/media/$disk/archive" 
+DISK="burak/New Volume"
+if [ $1 == "b" ]; then
+    DISK="burak/17EA-3758"
+fi
+echo "Copying files to /media/$DISK/archive"
+rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/kitaplar "/media/$DISK/archive"
+rsync  -W --size-only  --delete-after -P -avzh /home/burak/kod "/media/$DISK/archive" 
+rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/classnotes "/media/$DISK/archive" 
+rsync  -W --size-only  --delete-after -P -avzh /home/burak/Documents/emacs-ipython "/media/$DISK/archive" 
+rsync  -W --size-only  --delete-after -P -avzh /home/skorsky/mindmeld "/media/$DISK/archive" 
+rsync  -W --size-only  --delete-after -P -avzh /home/burak/Dropbox "/media/$DISK/archive" 
+#rsync  -W --size-only  --delete-after -P -avzh /home/burak/Music "/media/$DISK/archive" 

### 2016-01-04 site-lisp/deascify.py
-

@@ -35,14 +35,14 @@ def convert():
     content = content.replace(" ise","WWXXD7")
     content = content.replace("kontur","WWXXD8")
     content = content.replace("ODE","WWXXD9")
-    content = content.replace("olduguna","WWXXDx1")
     content = content.replace("Oklit","WWXXDx3")
     content = content.replace("karekok","WWXXDx4")
     content = content.replace("kismi","WWXXDx5")
-    content = content.replace("oldugunu","WWXXDx6")
     content = content.replace("integral","WWXXDx7")
     content = content.replace("acilimi","WWXXDx8")
     content = content.replace(" aci","WWXXDy1")
+    content = content.replace("gayri","WWXXDy2")
+    content = content.replace("oldugu","WWXXDy3")
 
     result = to_tr(content)
     result = result.replace("WWXXD1","verisi")
@@ -52,14 +52,14 @@ def convert():
     result = result.replace("WWXXD7"," ise" )
     result = result.replace("WWXXD8","kontur" )
     result = result.replace("WWXXD9","ODE" )
-    result = result.replace('WWXXDx1',u'olduğuna' )
     result = result.replace('WWXXDx3',u'Öklit' )
     result = result.replace('WWXXDx4',u'karekök' )
     result = result.replace('WWXXDx5',u'kısmi' )
-    result = result.replace('WWXXDx6',u'olduğunu' )
     result = result.replace('WWXXDx7',u'integral' )
     result = result.replace('WWXXDx8',u'acılımı' )
     result = result.replace('WWXXDy1',u' açı' )
+    result = result.replace('WWXXDy2',u' gayrı' )
+    result = result.replace('WWXXDy3',u'olduğu' )
     lisp.delete_region(block_begin, block_end)
     lisp.insert(result)
     lisp.goto_char(remember_where)

### 2016-01-05 find.py
-

@@ -1,0 +1,2 @@
+import sys
+if sys.argv[1] in open(sys.argv[2]).read(): print sys.argv[2]

### 2016-01-05 find.py
-

@@ -1,2 +1,5 @@
+# -*- coding: utf-8 -*-
+# Usage:
+# find . -name '*.tex' -exec python $HOME/kod/find.py '[string]'  {} \;
 import sys
 if sys.argv[1] in open(sys.argv[2]).read(): print sys.argv[2]

### 2016-01-08 books/Machine_Learning_Probabilistic_Perspective/dishonest.m
-

@@ -1,0 +1,8 @@
+
+prior_casino=[0.5;0.5];
+transmat_casino=[0.95 0.05;0.05 0.95];
+obsmat_casino=[1/6 1/6 1/6 1/6 1/6 1/6;1/10 1/10 1/10 1/10 1/10 1/2];
+data = [1 2 4 5 5 2 6 4 6 2 1 4 6 1 4 6 1 3 6 1 3 6 6 6 1 6 6 4 6 6 1 6 3 6 6 1 6 3 6 6 1 6 3 6 1 6 5 1 5 6 1 5 1 1 5 1 4 6 1 2 3 5 6 2 3 4 4];
+local=multinomial_prob(data,  obsmat_casino);
+path = viterbi_path(prior_casino, transmat_casino, local)
+

### 2016-01-08 books/Machine_Learning_Probabilistic_Perspective/dishonest.m
-

@@ -1,8 +1,0 @@
-
-prior_casino=[0.5;0.5];
-transmat_casino=[0.95 0.05;0.05 0.95];
-obsmat_casino=[1/6 1/6 1/6 1/6 1/6 1/6;1/10 1/10 1/10 1/10 1/10 1/2];
-data = [1 2 4 5 5 2 6 4 6 2 1 4 6 1 4 6 1 3 6 1 3 6 6 6 1 6 6 4 6 6 1 6 3 6 6 1 6 3 6 6 1 6 3 6 1 6 5 1 5 6 1 5 1 1 5 1 4 6 1 2 3 5 6 2 3 4 4];
-local=multinomial_prob(data,  obsmat_casino);
-path = viterbi_path(prior_casino, transmat_casino, local)
-

### 2016-01-08 books/Machine_Learning_Probabilistic_Perspective/dishon1.m
-

@@ -1,0 +1,8 @@
+% dishonest casino example 1 in matlab
+prior_casino=[0.5;0.5];
+transmat_casino=[0.95 0.05;0.05 0.95];
+obsmat_casino=[1/6 1/6 1/6 1/6 1/6 1/6;1/10 1/10 1/10 1/10 1/10 1/2];
+data = [1 2 4 5 5 2 6 4 6 2 1 4 6 1 4 6 1 3 6 1 3 6 6 6 1 6 6 4 6 6 1 6 3 6 6 1 6 3 6 6 1 6 3 6 1 6 5 1 5 6 1 5 1 1 5 1 4 6 1 2 3 5 6 2 3 4 4];
+local=multinomial_prob(data,  obsmat_casino);
+path = viterbi_path(prior_casino, transmat_casino, local)
+

### 2016-01-08 books/Machine_Learning_Probabilistic_Perspective/dishon2.m

@@ -1,0 +1,15 @@
+% dishonest casino example 2 in matlab
+
+data = [1 2 4 5 5 2 6 4 6 2 1 4 6 1 4 6 1 3 6 1 3 6 6 6 1 6 6 4 6 6 1 6 3 6 6 1 6 3 6 6 1 6 3 6 1 6 5 1 5 6 1 5 1 1 5 1 4 6 1 2 3 5 6 2 3 4 4];
+
+O = 6; Q = 2;
+% generate prior random matrices
+prior0 = normalise(rand(Q,1));
+transmat0 = mk_stochastic(rand(Q,Q));
+obsmat0 = mk_stochastic(rand(Q,O));
+
+[LL, prior1, transmat1, obsmat1] = dhmm_em(data, prior0, transmat0, obsmat0, 'max_iter', 10);
+
+prior1
+transmat1
+obsmat1

### 2016-01-09 books/Machine_Learning_Probabilistic_Perspective_Murphy/dishon2.m
-

@@ -3,6 +3,18 @@
 data = [1 2 4 5 5 2 6 4 6 2 1 4 6 1 4 6 1 3 6 1 3 6 6 6 1 6 6 4 6 6 1 6 3 6 6 1 6 3 6 6 1 6 3 6 1 6 5 1 5 6 1 5 1 1 5 1 4 6 1 2 3 5 6 2 3 4 4];
 
 O = 6; Q = 2;
+
+% generate prior random matrices
+prior0 = normalise(rand(Q,1));
+transmat0 = mk_stochastic(rand(Q,Q));
+obsmat0 = mk_stochastic(rand(Q,O));
+
+[LL, prior1, transmat1, obsmat1] = dhmm_em(data, prior0, transmat0, obsmat0, 'max_iter', 10);
+
+prior1
+transmat1
+obsmat1
+
 % generate prior random matrices
 prior0 = normalise(rand(Q,1));
 transmat0 = mk_stochastic(rand(Q,Q));

### 2016-01-10 site-lisp/deascify.py
-

@@ -43,6 +43,7 @@ def convert():
     content = content.replace(" aci","WWXXDy1")
     content = content.replace("gayri","WWXXDy2")
     content = content.replace("oldugu","WWXXDy3")
+    content = content.replace("Pandas","WWXXDy4")
 
     result = to_tr(content)
     result = result.replace("WWXXD1","verisi")
@@ -60,6 +61,7 @@ def convert():
     result = result.replace('WWXXDy1',u' açı' )
     result = result.replace('WWXXDy2',u' gayrı' )
     result = result.replace('WWXXDy3',u'olduğu' )
+    result = result.replace('WWXXDy4',u'Pandas' )
     lisp.delete_region(block_begin, block_end)
     lisp.insert(result)
     lisp.goto_char(remember_where)
